---
title: "A deep dive on Austria"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A deep dive on Austria}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

`vignette("euro-area-and-member-countries")` estimates eleven indices under one
common definition, which is what makes them comparable and also what stops any
one of them from being examined closely. This vignette does the opposite: one
country, taken apart.

Austria is the country to take apart, for two reasons that have nothing to do
with its size. It is one of only two regions in `eamd` that carry a national
block on top of the common columns -- house prices, a mortgage spread and two
credit-to-GDP ratios -- which are the closest thing in this data to the
`loanhpi` and `spread_30_mort` of Koop and Korobilis (2014), so Austria is where
the paper's own block composition can very nearly be reproduced. And its
financial history separates cleanly from the aggregate's: a severe 2008--09, a
mild sovereign crisis, and a recent episode the euro area as a whole did not
have.

```{r setup}
library("fincond")

set.seed(1234567)

data("eamd")

at <- eamd$at
```

## The Austrian block

```{r block}
core <- colnames(eamd$ea$financial)
national <- setdiff(colnames(at$financial), core)

national
c(financial = ncol(at$financial), quarters = nrow(at$macro))
```

Twenty-two common columns and `r length(national)` national ones. The common
block is described in `?eamd`; what matters here is that three of its columns --
`irate_short`, `spread_money` and `usd` -- are euro-area series sitting in an
Austrian block, because the money market and the euro's dollar rate are not
national quantities. Only `irate_long`, and through it `spread_term`, is
Austria's own among the rates.

The `r length(national)` national columns come from AustriaMacroData rather than
from EA-MD-QD, and they are appended after the common ones, so `at$financial[,
core]` recovers the strictly comparable set. Whether they earn their place is a
question this vignette answers rather than assumes.

## The index

```{r fit}
model <- create_fcimodel(y = at$macro, x = at$financial, p = 4, n = 1)
model <- add_priors(model)
model <- add_posterior_coefficients(model)

index <- -fci(model, sign_on = "esi")
```

The orientation is worth one paragraph, because Austria is the case where the
sibling vignette's caution does not bite. That vignette signs on `esi` and
negates, because the VIX loads weakly or with the wrong sign in several members
and so cannot anchor all eleven at once. In Austria it loads
`r round(colMeans(factor_loadings(model))[["vix"]], 2)` with the right sign, and
the paper's own convention returns exactly the same index:

```{r anchor}
round(colMeans(factor_loadings(model))[c("vix", "gpr", "esi")], 2)

identical(as.numeric(-fci(model, sign_on = "esi")),
          as.numeric(fci(model, sign_on = "vix")))
```

Signing on `esi` is kept below anyway, so that every number here can be read
against the eleven-region tables without a mental sign flip. But in Austria it
is a convenience rather than a necessity.

```{r plot-index}
plot(index, main = "Austrian financial conditions index", ylab = "Index",
     xlab = "")
abline(h = 0, lty = "dashed", col = "grey40")
```

```{r peaks}
as_quarter <- function(x) {
  yr <- floor(x + 1e-8)
  paste0(yr, " Q", round((x - yr) * 4) + 1)
}

ord <- order(-index)[1:8]
data.frame(quarter = as_quarter(as.numeric(time(index))[ord]),
           index = round(as.numeric(index)[ord], 2))
```

The peak is `r as_quarter(as.numeric(time(index))[which.max(index)])`, and the
quarters around it take three of the top four places. The exception is 2001 Q4,
the dot-com bust and the months after September 2001 -- a genuine tightening in
Austria, and the only pre-2008 quarter anywhere near the crisis readings.

The rest of the list is the surprise. Four of the eight tightest quarters in
twenty-six years are 2023 or later.

## Austria is not a small Germany

Austria's real economy is usually described through Germany, its largest
trading partner. Its financial conditions do not work that way.

```{r versus}
fit_region <- function(d) {
  m <- create_fcimodel(y = d$macro, x = d$financial, p = 4, n = 1)
  m <- add_priors(m)
  add_posterior_coefficients(m)
}

ea <- fit_region(eamd$ea)
de <- fit_region(eamd$de)

three <- ts.intersect(at = index,
                      ea = -fci(ea, sign_on = "esi"),
                      de = -fci(de, sign_on = "esi"))

round(cor(three), 2)
```

Austria tracks the euro area aggregate at
`r round(cor(three[, "at"], three[, "ea"]), 2)` and Germany at only
`r round(cor(three[, "at"], three[, "de"]), 2)`. That is not Austria being
unusual -- it is Germany, whose index is the least like the aggregate of the
eleven, for reasons the sibling vignette gives. Austria sits where a small,
open, bank-financed member of a currency union should sit: close to the common
cycle.

```{r plot-three}
ts.plot(three, col = c("firebrick", "grey40", "steelblue"),
        lwd = c(1.9, 1.3, 1.3), ylab = "Index", xlab = "")
abline(h = 0, lty = "dashed", col = "grey60")
legend("topleft", legend = c("Austria", "Euro area", "Germany"),
       col = c("firebrick", "grey40", "steelblue"), lwd = 1.6, bty = "n")
```

```{r episodes}
episode <- function(from, to) {
  round(colMeans(window(three, start = from, end = to)), 2)
}

t(rbind("2000-01 dot-com"    = episode(c(2000, 2), c(2001, 4)),
        "2006-07 boom"       = episode(c(2006, 1), c(2007, 4)),
        "2008-09 crisis"     = episode(c(2008, 4), c(2009, 3)),
        "2011-13 sovereign"  = episode(c(2011, 3), c(2013, 4)),
        "2020 pandemic"      = episode(c(2020, 1), c(2020, 4)),
        "2021 recovery"      = episode(c(2021, 1), c(2021, 4)),
        "2022 invasion"      = episode(c(2022, 1), c(2022, 4)),
        "2023-25 tightening" = episode(c(2023, 1), c(2025, 4))))
```

Two rows carry the Austrian story. In 2011--13 Austria averages
`r episode(c(2011, 3), c(2013, 4))[["at"]]` against the euro area's
`r episode(c(2011, 3), c(2013, 4))[["ea"]]`: the sovereign crisis was a
periphery event, and Austria was not in the periphery -- whatever its banks'
exposure to central and eastern Europe was thought to imply at the time.

And in 2023--25 the ordering reverses: Austria at
`r episode(c(2023, 1), c(2025, 4))[["at"]]`, the euro area at
`r episode(c(2023, 1), c(2025, 4))[["ea"]]`, Germany at
`r episode(c(2023, 1), c(2025, 4))[["de"]]`. That is the one episode in the
sample where Austria is decisively the tightest of the three, and it is not a
spike but a plateau.

## The recent tightening

```{r recent}
round(window(index, start = c(2022, 1)), 2)
```

The index crosses zero in 2022 Q3 and is above one in ten of the eleven
quarters from 2023 Q2. That is the longest stretch of tight readings in the
sample by some way -- the crisis managed three consecutive quarters above one,
2001 Q4 managed a single one -- and unlike 2008--09 there was no banking
failure to point at. So it is worth asking what the factor is actually reading.

```{r drivers}
standardised <- ts(scale(at$financial), start = start(at$financial),
                   frequency = 4)

recent <- colMeans(window(standardised, start = c(2023, 1), end = c(2025, 4)),
                   na.rm = TRUE)
round(sort(recent), 2)
```

Consumer confidence, industrial confidence and the headline sentiment index all
sit more than a standard deviation below their own averages, and so does
long-term household lending; household debt growth, both credit-to-GDP ratios
and house price growth follow. Construction confidence is the one survey that
does not, at `r round(recent[["conf_constr"]], 2)`. The tightening cycle shows
up not in the policy rate itself -- `irate_short` is a first difference, so a
completed cycle of hikes returns it to around zero -- but in what it did to
credit and to confidence.

The interesting non-driver is the term spread. It is the second most extreme
input in that table, at `r round(recent[["spread_term"]], 2)` standard
deviations: an inverted curve, of the kind that usually dominates a stress
index. It contributes almost nothing here, because its loading had already
gone:

```{r term-spread}
loadings <- factor_loadings(model)
oriented <- -sign(colMeans(loadings)[["esi"]]) * loadings

round(oriented[c(1, 40, 80, nrow(oriented)), "spread_term"], 2)
```

A constant-loading model would have had to carry the 2023 inversion at the
weight the series earned in 2000. This one had already discounted it.

## What moves and what does not

```{r loadings-plot, fig.height = 9}
plot(model, which = "loadings")
```

Twenty-six paths are more than can be read at once, so the compact version is
the average loading by era, signed the way the index is:

```{r loadings-era}
era <- function(from, to) {
  round(colMeans(window(oriented, start = from, end = to)), 2)
}

t(rbind("2000-07" = era(c(2000, 2), c(2007, 4)),
        "2008-12" = era(c(2008, 1), c(2012, 4)),
        "2013-19" = era(c(2013, 1), c(2019, 4)),
        "2023-25" = era(c(2023, 1), c(2025, 4))))
```

Half of these loadings are negative, so a signed difference would confuse a
negative loading growing with a positive one shrinking. The change in magnitude
is what says whether a series gained weight in the index:

```{r loadings-drift}
drift <- abs(oriented[nrow(oriented), ]) - abs(oriented[1, ])

round(sort(drift, decreasing = TRUE)[1:5], 2)
round(sort(drift)[1:5], 2)
```

The drift is modest everywhere -- this is a factor that has kept its shape --
but it is not zero, and it points one way. The survey balances gain: industrial
confidence by `r round(drift[["conf_ind"]], 2)` and `esi` by
`r round(drift[["esi"]], 2)`. Corporate credit gains with them. What loses
weight is household credit, the term spread and the dollar rate. Over
twenty-six years the Austrian factor became more of a
survey-and-corporate-credit factor and less of a household-credit-and-rates
one.

## What the index is made of

Loadings say what each series contributes on average. They do not say what the
index would look like without it, which is a different question when
twenty-six correlated series share one factor. Refitting without each series in
turn:

```{r drop-one}
drop_one <- sapply(colnames(at$financial), function(series) {
  keep <- setdiff(colnames(at$financial), series)
  m <- create_fcimodel(y = at$macro, x = at$financial[, keep], p = 4, n = 1)
  m <- add_posterior_coefficients(add_priors(m))
  anchor <- if ("esi" %in% keep) "esi" else "conf_ind"
  cor(as.numeric(-fci(m, sign_on = anchor)), as.numeric(index))
})

round(sort(drop_one)[1:8], 3)
```

No single series matters much: drop the most influential one, `esi`, and the
index still correlates `r round(min(drop_one), 2)` with the full one. That is
the reassuring answer and also the uninformative one, because twenty-six
correlated series can each be individually redundant while a whole group of
them is not. Dropping groups is the sharper test:

```{r drop-group}
groups <- list(
  surveys     = c("conf_cons", "conf_ind", "conf_constr", "esi"),
  credit      = c("hh_debt", "hh_loans_lt", "hh_loans_st", "nfc_debt",
                  "nfc_loans_lt", "bank_loans_lt", "credit_gdp_hh",
                  "credit_gdp_nfc"),
  prices      = c("equity", "house_prices", "reer", "usd"),
  rates       = c("irate_short", "irate_long", "spread_term", "spread_money",
                  "spread_mortgage"),
  uncertainty = c("gpr", "vix"),
  national    = national)

# The anchor has to survive the drop, so take the first available series that
# rises with loosening; if every one of them goes, anchor on gpr, which rises
# with stress, and do not negate.
orient <- function(m, keep) {
  loose <- intersect(c("esi", "conf_ind", "hh_debt", "nfc_assets"), keep)
  if (length(loose) > 0) -fci(m, sign_on = loose[1]) else fci(m, sign_on = "gpr")
}

summarise <- function(i) {
  data.frame(cor = round(cor(as.numeric(i), as.numeric(index)), 3),
             peak = as_quarter(as.numeric(time(i))[which.max(i)]),
             crisis = round(mean(window(i, start = c(2008, 4),
                                        end = c(2009, 3))), 2),
             recent = round(mean(window(i, start = c(2023, 1),
                                        end = c(2025, 4))), 2))
}

without <- function(series) {
  keep <- setdiff(colnames(at$financial), series)
  m <- create_fcimodel(y = at$macro, x = at$financial[, keep], p = 4, n = 1)
  summarise(orient(add_posterior_coefficients(add_priors(m)), keep))
}

dropped <- do.call(rbind, lapply(groups, without))
rbind(dropped, "(none)" = summarise(index))
```

Two groups are load-bearing and four are not.

Without the surveys the index is a different series: the correlation falls to
`r dropped["surveys", "cor"]`, the 2008--09 average more than halves, and the
peak moves to 2001 Q4. A factor extracted from Austrian balance sheets, rates
and asset prices alone does not identify the financial crisis as the country's
worst quarter. That is a real dependence, and it is the caveat to carry away
from this vignette: the Austrian index is to a substantial degree a confidence
index.

Without credit the peak survives but the recent episode largely does not --
2023--25 falls to `r dropped["credit", "recent"]` against a full-block
`r round(mean(window(index, start = c(2023, 1), end = c(2025, 4))), 2)`. The two
episodes are made of different things. 2008--09 is legible in surveys and asset
prices; 2023--25 is a credit event, and an index without a credit block would
have understated it.

Dropping the national block moves the correlation to
`r dropped["national", "cor"]`, which is the honest answer to the question the
first section left open: the four Austrian series do not change the index. They
are worth having as a diagnostic -- the house price and mortgage spread paths
are what let the previous section say the property correction is not what the
factor was reading -- but nobody should expect the index to move because they
are there.

## A second factor

```{r two-factor}
model2 <- create_fcimodel(y = at$macro, x = at$financial, p = 4, n = 2)
model2 <- add_posterior_coefficients(add_priors(model2))

first <- -fci(model2, factor = 1, sign_on = "esi")
second <- fci(model2, factor = 2)

round(c(first_vs_index = cor(first, index),
        second_vs_index = cor(second, index)), 3)

round(sort(abs(colMeans(factor_loadings(model2, factor = 2))),
           decreasing = TRUE)[1:6], 2)
```

The first factor is the one-factor index back again. The second is orthogonal
to it and loads on bank lending, equity, the VIX, the dollar and the effective
exchange rate -- a market-and-external component that the single-factor index
merges into the domestic credit cycle. Whether separating it is worth doing
depends on the use; for reproducing the index of Koop and Korobilis (2014),
`n = 1` is the specification, and nothing here suggests it loses the main
signal.

## Robustness

The forgetting factors are the estimator's only real tuning parameters, and the
sample is short enough that they deserve a check rather than a default.

```{r decay}
grid <- list(
  "Koop-Korobilis" = list(u = 0.96, v = 0.96, lambda = 0.99, beta = 0.99),
  "no drift"       = list(u = 0.96, v = 0.96, lambda = 1,    beta = 1),
  "fast loadings"  = list(u = 0.96, v = 0.96, lambda = 0.95, beta = 0.99),
  "slow variances" = list(u = 0.99, v = 0.99, lambda = 0.99, beta = 0.99))

do.call(rbind, lapply(grid, function(d) {
  m <- create_fcimodel(y = at$macro, x = at$financial, p = 4, n = 1)
  i <- -fci(add_posterior_coefficients(add_priors(m, decay = d)),
            sign_on = "esi")
  data.frame(cor = round(cor(as.numeric(i), as.numeric(index)), 3),
             peak = as_quarter(as.numeric(time(i))[which.max(i)]),
             peak_value = round(max(i), 2))
}))
```

Every setting returns the same peak and a correlation of at least 0.99 with the
baseline, including the one that switches time variation off entirely. The
Austrian index is not an artefact of the discounting.

## A recursive index

The paper's second exercise re-estimates the index once per real-time vintage.
`eamd` has no vintages, so what follows is the weaker version: an expanding
window that uses today's revised figures throughout. It answers whether the
index needs data from after the quarter it is dating. It says nothing about how
much revisions would move it.

```{r recursive}
ends <- as.numeric(time(window(at$macro, start = c(2006, 1))))
out <- numeric(length(ends))

for (i in seq_along(ends)) {
  yr <- floor(ends[i] + 1e-8)
  qtr <- round((ends[i] - yr) * 4) + 1

  x_i <- window(at$financial, end = c(yr, qtr))
  # Keep only the series that have started by this quarter.
  x_i <- x_i[, colSums(!is.na(x_i)) >= 2, drop = FALSE]

  m_i <- create_fcimodel(y = window(at$macro, end = c(yr, qtr)), x = x_i,
                         p = 4, n = 1)
  m_i <- add_posterior_coefficients(add_priors(m_i))

  # Not standardised: each fit has its own sample, so its own mean and
  # variance, and rescaling every one separately would put the readings on
  # eighty different scales.
  i_i <- -fci(m_i, sign_on = "esi", standardize = FALSE)
  out[i] <- as.numeric(i_i)[length(i_i)]
}

recursive <- ts(out, start = c(2006, 1), frequency = 4)
```

Each entry is a one-off reading: the value the index took in that quarter,
estimated with nothing later in the sample.

```{r recursive-compare}
full <- -fci(model, sign_on = "esi", standardize = FALSE)
compared <- ts.intersect(recursive, full)

round(cor(compared[, 1], compared[, 2]), 3)
```

```{r recursive-plot}
scale_it <- function(v) (v - mean(v)) / stats::sd(v)
plotted <- ts(cbind(scale_it(compared[, 1]), scale_it(compared[, 2])),
              start = start(compared), frequency = 4)

ts.plot(plotted, col = c("firebrick", "grey30"), lwd = c(1.8, 1),
        ylab = "Index", xlab = "",
        main = "Expanding window vs. full sample")
legend("topleft", legend = c("Expanding window", "Full sample"),
       col = c("firebrick", "grey30"), lwd = c(1.8, 1), bty = "n")
```

`r round(cor(compared[, 1], compared[, 2]), 3)` puts Austria near the top of the
eleven -- `vignette("euro-area-and-member-countries")` computes the full
ranking -- and the reason is visible in the episode table above rather than in
anything about Austrian data quality. Austria's largest event falls squarely
inside the recursion's own window, so even a model fitted only to data up to
2009 has already seen what calibrates the full-sample index. Greece sits at the
bottom of that ranking for the mirror-image reason: a recursion beginning in
2006 had not yet met the crisis that defines the Greek factor. Read this number
as a statement about where Austria's crisis sits in its sample, not as a
general claim about the estimator.

## What the index is for

An index of financial conditions should say something about the real economy
that the real data have not said yet.

```{r predictive}
predictive <- function(target, h) {
  d <- as.data.frame(na.omit(stats::ts.intersect(
    lead = stats::lag(at$macro[, target], h),
    own = at$macro[, target],
    fci = index)))
  own_only <- summary(stats::lm(lead ~ own, data = d))$r.squared
  with_fci <- summary(stats::lm(lead ~ own + fci, data = d))
  c(horizon = h,
    r2_own = round(own_only, 3),
    r2_with_fci = round(with_fci$r.squared, 3),
    coef_fci = round(stats::coef(with_fci)[3, 1], 2))
}

t(sapply(c(1, 2, 4, 8), predictive, target = "u"))
```

Tighter conditions today mean higher Austrian unemployment later, at every
horizon tried. The gain is largest where it is most useful: a year ahead, the
unemployment rate's own persistence explains
`r predictive("u", 4)[["r2_own"]]` of the variance and the index takes that to
`r predictive("u", 4)[["r2_with_fci"]]`. At two years the persistence is gone
altogether -- own-lag `r predictive("u", 8)[["r2_own"]]` -- and the index
accounts for nearly all of the very little that is left.

```{r predictive-y}
t(sapply(c(1, 2, 4), predictive, target = "y"))
```

For output growth the sign reverses, as it should, since growth is the outcome
you want more of: tighter conditions now, weaker growth one and two quarters
out. By a year ahead there is nothing left to find in either direction, and the
fits are small throughout because a quarterly growth rate is close to
unforecastable at any horizon by anything -- which is not a defect of the
index.

Two caveats, the same two the sibling vignette gives. These are in-sample fits,
not forecasts: nothing here was estimated without the data it is scored on. And
the horizons overlap, so the regressions violate the assumptions their standard
errors would need, which is why none are reported. Read the tables as
descriptions of co-movement, and see Koop and Korobilis (2014, sec. 5) for the
out-of-sample version of the question.

## References

Barigozzi, M., & Lissona, C. EA-MD-QD: Large euro area and euro member countries
datasets for macroeconomic research.
<https://github.com/BarigozziMatteo/EA-MD-QD>

Caldara, D., & Iacoviello, M. (2022). Measuring geopolitical risk. *American
Economic Review, 112*(4), 1194--1225.

Koop, G., & Korobilis, D. (2014). A new index of financial conditions. *European
Economic Review, 71*, 101--116.
