---
title: "A deep dive on Austria"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A deep dive on Austria}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

`vignette("euro-area-and-member-countries")` estimates eleven indices under one
common definition, which is what makes them comparable and also what stops any
one of them from being examined closely. This vignette does the opposite: one
country, taken apart.

Austria is the country to take apart, for three reasons that have nothing to do
with its size. It is one of only two regions in `eamd` that carry a national
block on top of the common columns -- house prices, a mortgage spread and two
credit-to-GDP ratios -- which are the closest thing in this data to the
`loanhpi` and `spread_30_mort` of Koop and Korobilis (2014), so Austria is where
the paper's own block composition can very nearly be reproduced. Its financial
history separates cleanly from the aggregate's: a severe 2008--09, a mild
sovereign crisis, and a recent episode the euro area as a whole did not have.
And it is one of the two countries whose national statistics reach back far
enough to escape `eamd`'s 2000 start altogether, which is what `atlong` and the
last third of this vignette are for -- the eleven-region panel cannot be
extended, but Austria on its own can.

```{r setup}
library("fincond")

set.seed(1234567)

data("eamd")

at <- eamd$at
```

## The Austrian block

```{r block}
core <- colnames(eamd$ea$financial)
national <- setdiff(colnames(at$financial), core)

national
c(financial = ncol(at$financial), quarters = nrow(at$macro))
```

Twenty-two common columns and `r length(national)` national ones. The common
block is described in `?eamd`; what matters here is that three of its columns --
`irate_short`, `spread_money` and `usd` -- are euro-area series sitting in an
Austrian block, because the money market and the euro's dollar rate are not
national quantities. Only `irate_long`, and through it `spread_term`, is
Austria's own among the rates.

The `r length(national)` national columns come from AustriaMacroData rather than
from EA-MD-QD, and they are appended after the common ones, so `at$financial[,
core]` recovers the strictly comparable set. Whether they earn their place is a
question this vignette answers rather than assumes.

## The index

```{r fit}
model <- create_fcimodel(y = at$macro, x = at$financial, p = 4, n = 1)
model <- add_priors(model)
model <- add_posterior_coefficients(model)

index <- fci(model, sign_on = "esi")
```

The orientation is worth a paragraph, because it is chosen rather than
estimated and because this vignette does not choose it the way its siblings do.
Here the index is signed on `esi`, the economic sentiment indicator, and left
alone: sentiment rises when conditions loosen, so **a negative reading is a
tight one** and a positive reading a loose one. Zero is the sample average, not
a neutral stance.

That is the opposite of the convention in
`vignette("euro-area-and-member-countries")` and in
`vignette("reproducing-koop-korobilis")`, which both follow Koop and Korobilis
(2014) in making the index rise with stress. Every number below therefore needs
a sign flip before it is read against the eleven-region tables. The two
orientations are otherwise the same series:

```{r anchor}
round(colMeans(factor_loadings(model))[c("vix", "gpr", "esi")], 2)

identical(as.numeric(fci(model, sign_on = "esi")),
          as.numeric(-fci(model, sign_on = "vix")))
```

Austria is a country where either anchor would have served. The VIX loads
`r round(colMeans(factor_loadings(model))[["vix"]], 2)` with the right sign
here, so `-fci(model, sign_on = "vix")` returns exactly the same index. The
sibling vignette cannot rely on that, because the VIX loads weakly or with the
wrong sign in several members and so cannot anchor all eleven at once; `esi` is
the anchor that works everywhere, and it is kept here too.

```{r plot-index}
plot(index, main = "Austrian financial conditions index", ylab = "Index",
     xlab = "")
abline(h = 0, lty = "dashed", col = "grey40")
```

```{r peaks}
as_quarter <- function(x) {
  yr <- floor(x + 1e-8)
  paste0(yr, " Q", round((x - yr) * 4) + 1)
}

ord <- order(index)[1:8]
data.frame(quarter = as_quarter(as.numeric(time(index))[ord]),
           index = round(as.numeric(index)[ord], 2))
```

The tightest quarter is
`r as_quarter(as.numeric(time(index))[which.min(index)])`, the most negative
reading in the sample, and the quarters around it take three of the four
lowest places. The exception is 2001 Q4, the dot-com bust and the months after
September 2001 -- a genuine tightening in Austria, and the only pre-2008
quarter anywhere near the crisis readings.

The rest of the list is the surprise. Four of the eight tightest quarters in
twenty-six years are 2023 or later.

## Austria is not a small Germany

Austria's real economy is usually described through Germany, its largest
trading partner. Its financial conditions do not work that way.

```{r versus}
fit_region <- function(d) {
  m <- create_fcimodel(y = d$macro, x = d$financial, p = 4, n = 1)
  m <- add_priors(m)
  add_posterior_coefficients(m)
}

ea <- fit_region(eamd$ea)
de <- fit_region(eamd$de)

three <- ts.intersect(at = index,
                      ea = fci(ea, sign_on = "esi"),
                      de = fci(de, sign_on = "esi"))

round(cor(three), 2)
```

Austria tracks the euro area aggregate at
`r round(cor(three[, "at"], three[, "ea"]), 2)` and Germany at only
`r round(cor(three[, "at"], three[, "de"]), 2)`. That is not Austria being
unusual -- it is Germany, whose index is the least like the aggregate of the
eleven, for reasons the sibling vignette gives. Austria sits where a small,
open, bank-financed member of a currency union should sit: close to the common
cycle.

```{r plot-three}
ts.plot(three, col = c("firebrick", "grey40", "steelblue"),
        lwd = c(1.9, 1.3, 1.3), ylab = "Index", xlab = "")
abline(h = 0, lty = "dashed", col = "grey60")
legend("topleft", legend = c("Austria", "Euro area", "Germany"),
       col = c("firebrick", "grey40", "steelblue"), lwd = 1.6, bty = "n")
```

```{r episodes}
episode <- function(from, to) {
  round(colMeans(window(three, start = from, end = to)), 2)
}

t(rbind("2000-01 dot-com"    = episode(c(2000, 2), c(2001, 4)),
        "2006-07 boom"       = episode(c(2006, 1), c(2007, 4)),
        "2008-09 crisis"     = episode(c(2008, 4), c(2009, 3)),
        "2011-13 sovereign"  = episode(c(2011, 3), c(2013, 4)),
        "2020 pandemic"      = episode(c(2020, 1), c(2020, 4)),
        "2021 recovery"      = episode(c(2021, 1), c(2021, 4)),
        "2022 invasion"      = episode(c(2022, 1), c(2022, 4)),
        "2023-25 tightening" = episode(c(2023, 1), c(2025, 4))))
```

Remember that a lower number is a tighter one. Two rows carry the Austrian
story. In 2011--13 Austria averages
`r episode(c(2011, 3), c(2013, 4))[["at"]]` against the euro area's
`r episode(c(2011, 3), c(2013, 4))[["ea"]]` -- much the milder of the two:
the sovereign crisis was a periphery event, and Austria was not in the
periphery, whatever its banks' exposure to central and eastern Europe was
thought to imply at the time.

And in 2023--25 the ordering reverses: Austria at
`r episode(c(2023, 1), c(2025, 4))[["at"]]`, the euro area at
`r episode(c(2023, 1), c(2025, 4))[["ea"]]`, Germany at
`r episode(c(2023, 1), c(2025, 4))[["de"]]`. Austria is now the most negative
of the three, which is the one episode in the sample where it is decisively
the tightest, and it is not a spike but a plateau.

## The recent tightening

```{r recent}
round(window(index, start = c(2022, 1)), 2)
```

The index crosses zero downwards in 2022 Q3 and is below minus one in ten of
the eleven quarters from 2023 Q2. That is the longest stretch of tight readings
in the sample by some way -- the crisis managed three consecutive quarters
below minus one, 2001 Q4 managed a single one -- and unlike 2008--09 there was
no banking failure to point at. So it is worth asking what the factor is
actually reading.

```{r drivers}
standardised <- ts(scale(at$financial), start = start(at$financial),
                   frequency = 4)

recent <- colMeans(window(standardised, start = c(2023, 1), end = c(2025, 4)),
                   na.rm = TRUE)
round(sort(recent), 2)
```

Consumer confidence, industrial confidence and the headline sentiment index all
sit more than a standard deviation below their own averages, and so does
long-term household lending; household debt growth, both credit-to-GDP ratios
and house price growth follow. Construction confidence is the one survey that
does not, at `r round(recent[["conf_constr"]], 2)`. The tightening cycle shows
up not in the policy rate itself -- `irate_short` is a first difference, so a
completed cycle of hikes returns it to around zero -- but in what it did to
credit and to confidence.

The interesting non-driver is the term spread. It is the second most extreme
input in that table, at `r round(recent[["spread_term"]], 2)` standard
deviations: an inverted curve, of the kind that usually dominates a stress
index. It contributes almost nothing here, because its loading had already
gone:

```{r term-spread}
loadings <- factor_loadings(model)
oriented <- sign(colMeans(loadings)[["esi"]]) * loadings

round(oriented[c(1, 40, 80, nrow(oriented)), "spread_term"], 2)
```

A constant-loading model would have had to carry the 2023 inversion at the
weight the series earned in 2000. This one had already discounted it.

## Reading the index by contribution

The table above ranks inputs by how far they sat from their own averages, which
says what was extreme rather than what the index was weighting. The term spread
has just shown the difference: `r round(abs(recent[["spread_term"]]), 2)`
standard deviations of inversion, the second largest deviation in the table,
and no effect. Weighting properly needs both quantities the filter itself uses -- the
loading, and how precisely that series is measured, which is what
`posterior$u_sigma` tracks. Weighting each observation that way and dividing by
the total precision gives a set of contributions that add up to the index:

```{r contributions}
contributions <- function(m) {
  l <- factor_loadings(m)
  v <- m$posterior$u_sigma
  orientation <- sign(colMeans(l)[["esi"]])

  # An absent series contributes nothing rather than propagating NA.
  obs <- m$data$x
  obs[is.na(obs)] <- 0

  # Precision weights: loading over measurement variance, normalised by the
  # total. This is the static generalised-least-squares view of the update
  # step, so it omits the term the filter takes from the transition equation.
  out <- ((orientation * l) / v) / rowSums((l^2) / v) * obs
  out <- out / stats::sd(orientation * as.numeric(m$posterior$factors[, 1]))

  colnames(out) <- colnames(m$data$x)
  stats::ts(out, start = stats::start(m$posterior$factors), frequency = 4)
}

contrib <- contributions(model)

# Worth checking before reading anything off it.
round(cor(rowSums(contrib), as.numeric(index)), 3)
```

`r round(cor(rowSums(contributions(model)), as.numeric(index)), 3)` is close
enough to read shares from, and not so close that the decimals should be
trusted; the omitted transition term is what the gap is made of.

```{r contributions-2022}
moved <- function(cb, from, to) {
  round(colMeans(window(cb, start = from, end = to)) -
          colMeans(window(cb, start = c(2021, 1), end = c(2021, 4))), 2)
}

y2022 <- moved(contrib, c(2022, 1), c(2022, 4))

sort(y2022)[1:8]
sort(y2022, decreasing = TRUE)[1:4]
```

Contributions carry the sign of the index, so a tightening contribution is a
negative number and the first block above is the tightening one.

2022 is the year Austria's index crossed from very loose to neutral -- the 2021
average is `r round(mean(window(index, start = c(2021, 1), end = c(2021, 4))), 2)`
and the 2022 average `r round(mean(window(index, start = c(2022, 1), end = c(2022, 4))), 2)`
-- and roughly half of that move is the survey balances. Consumer confidence
alone fell about three standard deviations over the year. Geopolitical risk
contributes `r y2022[["gpr"]]`, which is the invasion of Ukraine entering an
Austrian index through a global series, and corporate balance sheets most of
the rest.

The instructive entries are the ones in the second block, which pulled the
other way. The short rate *offsets* `r abs(y2022[["irate_short"]])` of the 2022
tightening, and the euro area's own index behaves the same way:

```{r contributions-ea}
ea_2022 <- moved(contributions(ea), c(2022, 1), c(2022, 4))

sort(ea_2022)[1:5]
sort(ea_2022, decreasing = TRUE)[1:3]
```

So the largest tightening cycle in the ECB's history enters both indices as a
*loosening* contribution. That follows from the negative loading the sibling
vignette explains: historically the ECB has cut into stress, so a rising policy
rate is what accompanies good conditions. In 2022 that relationship inverted --
hikes into a supply shock -- and a factor model estimated on the previous two
decades has no way to know. The reading for 2022 is, if anything, too loose,
and this is the clearest example in the vignette of a loading that is correct
about the sample and wrong about the moment.

The same decomposition puts the earlier drop-group result on a firmer footing.
Against the same 2021 base, the recent episode is built differently:

```{r contributions-recent}
sort(moved(contrib, c(2023, 1), c(2025, 4)))[1:8]
```

Household credit now sits among the largest contributors, where in 2022 it
barely registered: household debt contributed `r y2022[["hh_debt"]]` and
long-term household lending `r y2022[["hh_loans_lt"]]`, against
`r moved(contrib, c(2023, 1), c(2025, 4))[["hh_debt"]]` and
`r moved(contrib, c(2023, 1), c(2025, 4))[["hh_loans_lt"]]` now -- four to five
times the weight. That is the conclusion the group-dropping exercise reached,
2023--25 being a credit event where 2022 was a confidence one, arrived at from
the loadings rather than from refitting.

## What moves and what does not

```{r loadings-plot, fig.height = 9}
plot(model, which = "loadings")
```

Twenty-six paths are more than can be read at once, so the compact version is
the average loading by era, signed the way the index is:

```{r loadings-era}
era <- function(from, to) {
  round(colMeans(window(oriented, start = from, end = to)), 2)
}

t(rbind("2000-07" = era(c(2000, 2), c(2007, 4)),
        "2008-12" = era(c(2008, 1), c(2012, 4)),
        "2013-19" = era(c(2013, 1), c(2019, 4)),
        "2023-25" = era(c(2023, 1), c(2025, 4))))
```

Seven of these loadings are negative, so a signed difference would confuse a
negative loading growing with a positive one shrinking. The change in magnitude
is what says whether a series gained weight in the index:

```{r loadings-drift}
drift <- abs(oriented[nrow(oriented), ]) - abs(oriented[1, ])

round(sort(drift, decreasing = TRUE)[1:5], 2)
round(sort(drift)[1:5], 2)
```

The drift is modest everywhere -- this is a factor that has kept its shape --
but it is not zero, and it points one way. The survey balances gain: industrial
confidence by `r round(drift[["conf_ind"]], 2)` and `esi` by
`r round(drift[["esi"]], 2)`. Corporate credit gains with them. What loses
weight is household credit, the term spread and the dollar rate. Over
twenty-six years the Austrian factor became more of a
survey-and-corporate-credit factor and less of a household-credit-and-rates
one.

## What the index is made of

Loadings say what each series contributes on average. They do not say what the
index would look like without it, which is a different question when
twenty-six correlated series share one factor. Refitting without each series in
turn:

```{r drop-one}
drop_one <- sapply(colnames(at$financial), function(series) {
  keep <- setdiff(colnames(at$financial), series)
  m <- create_fcimodel(y = at$macro, x = at$financial[, keep], p = 4, n = 1)
  m <- add_posterior_coefficients(add_priors(m))
  anchor <- if ("esi" %in% keep) "esi" else "conf_ind"
  cor(as.numeric(fci(m, sign_on = anchor)), as.numeric(index))
})

round(sort(drop_one)[1:8], 3)
```

No single series matters much: drop the most influential one, `esi`, and the
index still correlates `r round(min(drop_one), 2)` with the full one. That is
the reassuring answer and also the uninformative one, because twenty-six
correlated series can each be individually redundant while a whole group of
them is not. Dropping groups is the sharper test:

```{r drop-group}
groups <- list(
  surveys     = c("conf_cons", "conf_ind", "conf_constr", "esi"),
  credit      = c("hh_debt", "hh_loans_lt", "hh_loans_st", "nfc_debt",
                  "nfc_loans_lt", "bank_loans_lt", "credit_gdp_hh",
                  "credit_gdp_nfc"),
  prices      = c("equity", "house_prices", "reer", "usd"),
  rates       = c("irate_short", "irate_long", "spread_term", "spread_money",
                  "spread_mortgage"),
  uncertainty = c("gpr", "vix"),
  national    = national)

# The anchor has to survive the drop, so take the first available series that
# rises with loosening and sign on it directly; if every one of them goes,
# anchor on gpr, which rises with stress, and negate.
orient <- function(m, keep) {
  loose <- intersect(c("esi", "conf_ind", "hh_debt", "nfc_assets"), keep)
  if (length(loose) > 0) fci(m, sign_on = loose[1]) else -fci(m, sign_on = "gpr")
}

summarise <- function(i) {
  data.frame(cor = round(cor(as.numeric(i), as.numeric(index)), 3),
             trough = as_quarter(as.numeric(time(i))[which.min(i)]),
             crisis = round(mean(window(i, start = c(2008, 4),
                                        end = c(2009, 3))), 2),
             recent = round(mean(window(i, start = c(2023, 1),
                                        end = c(2025, 4))), 2))
}

without <- function(series) {
  keep <- setdiff(colnames(at$financial), series)
  m <- create_fcimodel(y = at$macro, x = at$financial[, keep], p = 4, n = 1)
  summarise(orient(add_posterior_coefficients(add_priors(m)), keep))
}

dropped <- do.call(rbind, lapply(groups, without))
rbind(dropped, "(none)" = summarise(index))
```

Two groups are load-bearing and four are not.

Without the surveys the index is a different series: the correlation falls to
`r dropped["surveys", "cor"]`, the 2008--09 average loses more than half its
depth, and the trough moves to 2001 Q4. A factor extracted from Austrian
balance sheets, rates and asset prices alone does not identify the financial
crisis as the country's worst quarter. That is a real dependence, and it is the caveat to carry away
from this vignette: the Austrian index is to a substantial degree a confidence
index.

Without credit the trough survives but the recent episode largely does not --
2023--25 rises to `r dropped["credit", "recent"]` against a full-block
`r round(mean(window(index, start = c(2023, 1), end = c(2025, 4))), 2)`. The two
episodes are made of different things. 2008--09 is legible in surveys and asset
prices; 2023--25 is a credit event, and an index without a credit block would
have understated it.

Dropping the national block moves the correlation to
`r dropped["national", "cor"]`, which is the honest answer to the question the
first section left open: the four Austrian series do not change the index. They
are worth having as a diagnostic -- the house price and mortgage spread paths
are what let the contribution section say the property correction is not what
the factor was reading -- but nobody should expect the index to move because they
are there.

## Are the surveys just the cycle?

"To a substantial degree a confidence index" invites a sharper objection than
the drop-group table can answer. Survey balances move with the business cycle,
and Austria's four are general economic sentiment rather than anything
financial -- the euro area has no counterpart to the senior loan officer survey
that supplies `stdscom` in Koop and Korobilis's own block. If the surveys are
simply reporting current growth and unemployment back to the model, then what
the index calls financial conditions is partly the real economy wearing a
different label.

Some of that is answered by construction. The macro block is not an input the
factor has to compete with, it is a regressor in the same measurement equation:
each financial series is written as a loading on `y`, `p` and `u` plus a
loading on the factor, so the factor is already what the surveys do *beyond*
the macro block. The question is whether anything survives once lags are
allowed too.

```{r survey-macro-r2}
surveys <- groups$surveys

lg <- function(z, k) stats::lag(z, -k)
macro_now <- stats::ts.union(y = at$macro[, "y"], p = at$macro[, "p"],
                             u = at$macro[, "u"])
macro_lags <- stats::ts.union(
  macro_now,
  y1 = lg(at$macro[, "y"], 1), p1 = lg(at$macro[, "p"], 1),
  u1 = lg(at$macro[, "u"], 1),
  y4 = lg(at$macro[, "y"], 4), p4 = lg(at$macro[, "p"], 4),
  u4 = lg(at$macro[, "u"], 4))

on_macro <- function(series, regs) {
  d <- as.data.frame(stats::ts.intersect(s = at$financial[, series], regs))
  stats::lm(s ~ ., data = d[stats::complete.cases(d), , drop = FALSE])
}

round(t(sapply(surveys, function(s) {
  c(contemp = summary(on_macro(s, macro_now))$r.squared,
    with_lags = summary(on_macro(s, macro_lags))$r.squared)
})), 3)
```

Contemporaneously the macro block accounts for well under a fifth of each
survey. Allow a year of lags and that rises to between a third and three
fifths, which is enough to take the objection seriously. So take it seriously:
replace each survey by what the macro block cannot explain, leave it in its
place in the block, and refit.

```{r survey-residual-index}
residualise <- function(regs) {
  out <- at$financial
  stamp <- round(as.numeric(stats::time(out)), 6)
  for (s in surveys) {
    j <- stats::ts.intersect(s = at$financial[, s], regs)
    d <- as.data.frame(j)
    ok <- stats::complete.cases(d)
    e <- stats::residuals(stats::lm(s ~ ., data = d[ok, , drop = FALSE]))
    v <- rep(NA_real_, nrow(out))
    v[match(round(as.numeric(stats::time(j))[ok], 6), stamp)] <- e
    out[, s] <- v
  }
  out
}

refit <- function(xx) {
  m <- create_fcimodel(y = at$macro, x = xx, p = 4, n = 1)
  fci(add_posterior_coefficients(add_priors(m)), sign_on = "esi")
}

rbind("(none)"              = summarise(index),
      "purged of macro"     = summarise(refit(residualise(macro_now))),
      "purged, four lags"   = summarise(refit(residualise(macro_lags))),
      "surveys dropped"     = dropped["surveys", ])
```

The surveys are not a business-cycle proxy in disguise. Purging the
contemporaneous macro block changes nothing worth reporting, which is what the
measurement equation already implied. Purging a year of lags as well costs
something real -- the correlation falls and the crisis average softens -- but
the index still dates Austria's worst quarter to 2009 Q1. Only dropping the
surveys outright moves it to 2001 Q4.

So the dependence the previous section found is on survey information that `y`,
`p` and `u` do not contain at any of these lags, which is the version of the
claim worth keeping. What the surveys are *not* is a substitute for the credit
standards series the paper had and this block does not; that remains a
composition gap, and it is an argument about what Austria is missing rather
than about what its sentiment balances are made of.

One limit. The regressions above are fitted on the whole sample, so the purged
blocks use information no contemporaneous user would have had. This is a
description of what the index is made of, not a claim about what it could have
been in real time.

## A second factor

```{r two-factor}
model2 <- create_fcimodel(y = at$macro, x = at$financial, p = 4, n = 2)
model2 <- add_posterior_coefficients(add_priors(model2))

first <- fci(model2, factor = 1, sign_on = "esi")
second <- fci(model2, factor = 2)

round(c(first_vs_index = cor(first, index),
        second_vs_index = cor(second, index)), 3)

round(sort(abs(colMeans(factor_loadings(model2, factor = 2))),
           decreasing = TRUE)[1:6], 2)
```

The first factor is the one-factor index back again. The second is orthogonal
to it and loads on bank lending, equity, the VIX, the dollar and the effective
exchange rate -- a market-and-external component that the single-factor index
merges into the domestic credit cycle. Whether separating it is worth doing
depends on the use; for reproducing the index of Koop and Korobilis (2014),
`n = 1` is the specification, and nothing here suggests it loses the main
signal.

## Robustness

The forgetting factors are the estimator's only real tuning parameters, and the
sample is short enough that they deserve a check rather than a default.

```{r decay}
grid <- list(
  "Koop-Korobilis" = list(u = 0.96, v = 0.96, lambda = 0.99, beta = 0.99),
  "no drift"       = list(u = 0.96, v = 0.96, lambda = 1,    beta = 1),
  "fast loadings"  = list(u = 0.96, v = 0.96, lambda = 0.95, beta = 0.99),
  "slow variances" = list(u = 0.99, v = 0.99, lambda = 0.99, beta = 0.99))

do.call(rbind, lapply(grid, function(d) {
  m <- create_fcimodel(y = at$macro, x = at$financial, p = 4, n = 1)
  i <- fci(add_posterior_coefficients(add_priors(m, decay = d)),
            sign_on = "esi")
  data.frame(cor = round(cor(as.numeric(i), as.numeric(index)), 3),
             trough = as_quarter(as.numeric(time(i))[which.min(i)]),
             trough_value = round(min(i), 2))
}))
```

Every setting returns the same trough and a correlation of at least 0.99 with
the baseline, including the one that switches time variation off entirely. The
Austrian index is not an artefact of the discounting.

## A recursive index

The paper's second exercise re-estimates the index once per real-time vintage.
`eamd` has no vintages, so what follows is the weaker version: an expanding
window that uses today's revised figures throughout. It answers whether the
index needs data from after the quarter it is dating. It says nothing about how
much revisions would move it.

```{r recursive}
# Written as a function of the dataset because the longer sample below is put
# through exactly the same recursion. The whole filtered path is kept rather
# than only its last value: the forecasting section needs the index history
# each vintage would have had, not just its final reading.
vintages_of <- function(d, from) {
  origins <- as.numeric(time(window(d$macro, start = from)))

  paths <- lapply(origins, function(o) {
    yr <- floor(o + 1e-8)
    qtr <- round((o - yr) * 4) + 1

    x_o <- window(d$financial, end = c(yr, qtr))
    # Keep only the series that have started by this quarter.
    x_o <- x_o[, colSums(!is.na(x_o)) >= 2, drop = FALSE]

    m_o <- create_fcimodel(y = window(d$macro, end = c(yr, qtr)), x = x_o,
                           p = 4, n = 1)
    m_o <- add_posterior_coefficients(add_priors(m_o))

    # Not standardised: each fit has its own sample, so its own mean and
    # variance, and rescaling every one separately would put the readings on
    # eighty different scales.
    fci(m_o, sign_on = "esi", standardize = FALSE)
  })

  list(origins = origins, vintage = paths, macro = d$macro)
}

vintages <- vintages_of(at, from = c(2006, 1))

last_of <- function(z) as.numeric(z)[length(z)]
recursive <- ts(vapply(vintages$vintage, last_of, numeric(1)),
                start = c(2006, 1), frequency = 4)
```

Each entry of `recursive` is a one-off reading: the value the index took in that
quarter, estimated with nothing later in the sample.

```{r recursive-compare}
full <- fci(model, sign_on = "esi", standardize = FALSE)
compared <- ts.intersect(recursive, full)

round(cor(compared[, 1], compared[, 2]), 3)
```

```{r recursive-plot}
scale_it <- function(v) (v - mean(v)) / stats::sd(v)
plotted <- ts(cbind(scale_it(compared[, 1]), scale_it(compared[, 2])),
              start = start(compared), frequency = 4)

ts.plot(plotted, col = c("firebrick", "grey30"), lwd = c(1.8, 1),
        ylab = "Index", xlab = "",
        main = "Expanding window vs. full sample")
legend("topleft", legend = c("Expanding window", "Full sample"),
       col = c("firebrick", "grey30"), lwd = c(1.8, 1), bty = "n")
```

`r round(cor(compared[, 1], compared[, 2]), 3)` puts Austria near the top of the
eleven -- `vignette("euro-area-and-member-countries")` computes the full
ranking -- and the reason is visible in the episode table above rather than in
anything about Austrian data quality. Austria's largest event falls squarely
inside the recursion's own window, so even a model fitted only to data up to
2009 has already seen what calibrates the full-sample index. Greece sits at the
bottom of that ranking for the mirror-image reason: a recursion beginning in
2006 had not yet met the crisis that defines the Greek factor. Read this number
as a statement about where Austria's crisis sits in its sample, not as a
general claim about the estimator.

## A longer sample

Everything above runs on `eamd`, which starts in 2000 Q1 because EA-MD-QD
does -- for all eleven regions at once. That is a property of the source rather
than of Austria, and it is the binding constraint on a vignette that has just
spent a section complaining about an eighty-quarter evaluation sample. Austria
is one of the two regions where it can be relaxed, because AustriaMacroData
publishes a national panel reaching back to the 1950s. `atlong` is that panel
assembled to the same design:

```{r atlong-data}
data("atlong")

c(quarters = nrow(atlong$macro), series = ncol(atlong$financial))
range(time(atlong$macro))
```

`r nrow(atlong$macro)` quarters against `r nrow(at$macro)`. What stops it going
further is the unemployment rate, which neither source carries before 1993 in a
form that also reaches the present; real GDP would allow 1960. The price of the
extra years is that the block is ragged, since the series do not all begin at
once:

```{r atlong-ragged}
# Counted over the span the model actually uses: the financial block runs a few
# quarters past the macro one, and those quarters are dropped on alignment.
covered <- window(atlong$financial, start = start(atlong$macro),
                  end = end(atlong$macro))
available <- rowSums(!is.na(covered))

c(first = available[1], last = available[length(available)])

# How many quarters the block has each width for.
table(available)
```

That is the shape `koop` has -- its VIX starts in 1990 against a sample from
1970 -- and what `create_fcimodel` carries rather than imputes. The sectoral
balance sheets arrive last, in 2000, because no national source publishes them.

```{r atlong-fit}
long_model <- create_fcimodel(y = atlong$macro, x = atlong$financial,
                              p = 4, n = 1)
long_model <- add_posterior_coefficients(add_priors(long_model))

long_index <- fci(long_model, sign_on = "esi")
```

The first thing to ask of a differently-sourced index is whether it agrees with
the one it extends, over the quarters both cover:

```{r atlong-agree}
overlap <- ts.intersect(long = long_index, eamd = index)

round(cor(overlap[, "long"], overlap[, "eamd"]), 3)
```

It does, at `r round(cor(ts.intersect(long_index, index)[, 1], ts.intersect(long_index, index)[, 2]), 3)`,
which is worth more than it looks. The two blocks share no interest rate
construction, take equity and house prices from different publishers, and use
different inflation measures -- `atlong` has the consumer price index because
the national panel publishes no deflator. Two independent assemblies of the
idea "Austrian financial conditions" landing on the same series is evidence
about the idea rather than about either dataset.

```{r atlong-plot}
ts.plot(long_index, col = "firebrick", lwd = 1.6, ylab = "Index", xlab = "",
        main = "Austrian financial conditions, 1993 to 2025")
abline(h = 0, lty = "dashed", col = "grey40")
```

```{r atlong-peaks}
ord <- order(long_index)[1:8]
data.frame(quarter = as_quarter(as.numeric(time(long_index))[ord]),
           index = round(as.numeric(long_index)[ord], 2))
```

The ranking survives the extension -- `r as_quarter(as.numeric(time(long_index))[which.min(long_index)])`
is still the trough, and 2001 Q4 still second -- and the new years earn a place
in it rather than sitting quietly at the start. 1993 Q1 and 1993 Q2 come third
and sixth:

```{r atlong-1993}
round(window(long_index, start = c(1993, 1), end = c(1995, 4)), 2)
```

Nothing in the estimation was told about 1993, and it is the right answer: the
exchange rate mechanism crisis ran from September 1992 into August 1993, and
Austria was in recession through it, three years short of joining the European
Union and six from the euro. An index that reads it as the tightest stretch
outside 2008 and 2001 is making a defensible claim about a period no part of
`eamd` can see.

Two cautions before using this in place of `eamd`. It is one country, so
nothing here is comparable with the other ten -- that is exactly the trade the
dataset makes, and `?atlong` states it. And the early quarters rest on eleven
series rather than twenty-seven, so the 1990s readings are drawn from surveys,
rates and asset prices with no balance-sheet information at all; the credit
block that the drop-group section found load-bearing for 2023--25 simply does
not exist before 2000.

## What the index is for

An index of financial conditions should say something about the real economy
that the real data have not said yet.

```{r predictive}
predictive <- function(target, h) {
  d <- as.data.frame(na.omit(stats::ts.intersect(
    lead = stats::lag(at$macro[, target], h),
    own = at$macro[, target],
    fci = index)))
  own_only <- summary(stats::lm(lead ~ own, data = d))$r.squared
  with_fci <- summary(stats::lm(lead ~ own + fci, data = d))
  c(horizon = h,
    r2_own = round(own_only, 3),
    r2_with_fci = round(with_fci$r.squared, 3),
    coef_fci = round(stats::coef(with_fci)[3, 1], 2))
}

t(sapply(c(1, 2, 4, 8), predictive, target = "u"))
```

Tighter conditions today mean higher Austrian unemployment later, at every
horizon tried. Because tight is negative here, that relationship shows up as a
*negative* `coef_fci`: a fall in the index is followed by a rise in
unemployment. The gain is largest where it is most useful: a year ahead, the
unemployment rate's own persistence explains
`r predictive("u", 4)[["r2_own"]]` of the variance and the index takes that to
`r predictive("u", 4)[["r2_with_fci"]]`. At two years the persistence is gone
altogether -- own-lag `r predictive("u", 8)[["r2_own"]]` -- and the index
accounts for nearly all of the very little that is left.

```{r predictive-y}
t(sapply(c(1, 2, 4), predictive, target = "y"))
```

For output growth the coefficient turns positive, which is the same economics
read through the same orientation: growth is the outcome you want more of, so a
lower index -- tighter conditions now -- goes with weaker growth one and two
quarters out. By a year ahead there is nothing left to find in either
direction, and the fits are small throughout because a quarterly growth rate is
close to
unforecastable at any horizon by anything -- which is not a defect of the
index.

Both tables describe co-movement rather than forecasting skill: nothing in them
was estimated without the data it is scored on, and the horizons overlap, so
the standard errors those regressions would need do not hold. The next section
does the version that is not vulnerable to either objection.

## Out of sample

The vintages built for the recursive index make a real forecasting exercise
possible,
because they contain the index history a user would have had at each date. The
rule is that nothing estimated at origin `t` may touch data after `t` -- which
covers the regression coefficients as well as the index, since a regression
fitted on the full sample would leak the future just as surely as a full-sample
factor would.

```{r oos-forecast}
forecasts <- function(target, h, set = vintages) {
  y <- set$macro[, target]
  stamp <- as.numeric(time(y))

  rows <- lapply(seq_along(set$origins), function(k) {
    path <- set$vintage[[k]]
    realised <- which(abs(stamp - (set$origins[k] + h / 4)) < 1e-6)
    if (length(realised) == 0) return(NULL)

    # Everything below sees only data through this origin.
    seen <- window(y, end = end(path))
    d <- as.data.frame(na.omit(stats::ts.intersect(
      lead = stats::lag(seen, h), own = seen, fci = path)))
    if (nrow(d) < 12) return(NULL)

    now <- data.frame(own = last_of(seen), fci = last_of(path))
    predict_with <- function(f) {
      as.numeric(stats::predict(stats::lm(f, data = d), newdata = now))
    }

    data.frame(origin = set$origins[k], actual = as.numeric(y)[realised],
               bench = predict_with(lead ~ own),
               fci = predict_with(lead ~ own + fci))
  })

  do.call(rbind, rows)
}
```

The benchmark is the same own-lag regression the in-sample tables used, so the
comparison isolates what the index adds. Two statistics summarise a set of
forecasts. The ratio of root mean squared errors is what a user would have
experienced. The second needs a word of explanation:

```{r oos-score}
# Long-run variance, since h-step forecasts overlap for h > 1.
hac <- function(z, lag) {
  n <- length(z)
  z <- z - mean(z)
  v <- sum(z^2) / n
  for (l in seq_len(lag)) {
    v <- v + 2 * (1 - l / (lag + 1)) * sum(z[(l + 1):n] * z[1:(n - l)]) / n
  }
  v
}

score <- function(target, h, set = vintages) {
  f <- forecasts(target, h, set)
  e_bench <- f$actual - f$bench
  e_fci <- f$actual - f$fci

  # Clark and West (2007). A plain Diebold-Mariano test is undersized when the
  # models are nested, as these are: under the null the larger model estimates
  # a coefficient that is truly zero, and the noise in doing so inflates its
  # squared error. The last term subtracts exactly that penalty.
  adjusted <- e_bench^2 - (e_fci^2 - (f$bench - f$fci)^2)

  c(h = h, n = nrow(f),
    rmse_bench = round(sqrt(mean(e_bench^2)), 3),
    rmse_fci = round(sqrt(mean(e_fci^2)), 3),
    ratio = round(sqrt(mean(e_fci^2)) / sqrt(mean(e_bench^2)), 3),
    cw = round(mean(adjusted) /
                 sqrt(hac(adjusted, max(h - 1, 0)) / length(adjusted)), 2))
}
```

```{r oos-u}
t(sapply(c(1, 2, 4, 8), score, target = "u"))
```

A ratio below one is the index winning. For unemployment it wins at one, two
and four quarters, by up to `r round(100 * (1 - score("u", 2)[["ratio"]]))`
per cent of root mean squared error at two quarters, with Clark-West
statistics around three. At two years it loses, and the statistic is
`r score("u", 8)[["cw"]]` -- no evidence of anything.

The four-quarter row is the one to read carefully, because the two columns
disagree: a `r score("u", 4)[["ratio"]]` ratio is a small gain, while a
Clark-West statistic of `r score("u", 4)[["cw"]]` is a confident one. They are
answering different questions. The statistic asks whether the index belongs in
the population regression, and corrects for the estimation noise a user
actually has to pay; the ratio reports what that user got after paying it. At
one and two quarters the signal is large enough to clear its own estimation
cost. At four it is real but barely does.

```{r oos-y}
t(sapply(c(1, 2, 4), score, target = "y"))
t(sapply(c(1, 2, 4), score, target = "p"))
```

For output growth and inflation there is nothing to speak of. Two of the six
ratios dip below one, both by less than a point and a half, and the largest
Clark-West statistic is `r score("p", 4)[["cw"]]`, for inflation a year out --
which would be marginal on its own and is one row in six. The in-sample
table's
`r predictive("y", 1)[["r2_own"]]` to `r predictive("y", 1)[["r2_with_fci"]]`
gain in output-growth fit does not survive being asked to forecast, which is
the ordinary fate of in-sample gains that size and the reason this section
exists.

A gain concentrated in two crises would be a different claim from a gain earned
throughout, so it is worth splitting:

```{r oos-episodes}
f <- forecasts("u", 2)
gain <- (f$actual - f$bench)^2 - (f$actual - f$fci)^2

crisis <- f$origin >= 2008 & f$origin < 2010
pandemic <- f$origin >= 2020 & f$origin < 2021
calm <- !crisis & !pandemic

round(c(share_2008_09 = sum(gain[crisis]) / sum(gain),
        share_2020 = sum(gain[pandemic]) / sum(gain)), 2)

round(c(n = sum(calm),
        ratio_excluding_both = sqrt(mean((f$actual - f$fci)[calm]^2)) /
          sqrt(mean((f$actual - f$bench)[calm]^2))), 3)
```

It is not a crisis artefact. The 2008--09 origins contribute
`r round(100 * sum(gain[crisis]) / sum(gain))` per cent of the total gain and
the 2020 origins `r round(100 * sum(gain[pandemic]) / sum(gain))` per cent,
between them about a quarter of it from about a sixth of the sample. Drop both
and the ratio over the remaining `r sum(calm)` quarters is
`r round(sqrt(mean((f$actual - f$fci)[calm]^2)) / sqrt(mean((f$actual - f$bench)[calm]^2)), 3)`
-- slightly better than the full-sample figure, not worse.

```{r oos-plot}
ts.plot(ts(cumsum(gain), start = c(2006, 1), frequency = 4),
        col = "firebrick", lwd = 1.8, ylab = "Cumulative squared-error gain",
        xlab = "", main = "Where the forecast gain accrues (u, h = 2)")
abline(h = 0, lty = "dashed", col = "grey40")
```

The cumulative line is the honest summary of the exercise. A line that jumped
once and then flattened would mean an index that called one crisis and was
otherwise idle. This one climbs through the calm years as well, which is the
stronger result and the harder one to get.

Eighty forecasts of one country is a thin basis for any of this, and the
longer sample is the obvious thing to check it against. `atlong` reaches back
to 1993, so the same recursion started after an equivalent warm-up runs from
1999 and yields around a third more origins -- and, more usefully, ones drawn
from a different decade:

```{r oos-long}
long_vintages <- vintages_of(atlong, from = c(1999, 1))

length(long_vintages$origins)

t(sapply(c(1, 2, 4, 8), score, target = "u", set = long_vintages))
```

The result holds. Every sign is the same, the ratio still bottoms at two
quarters, the Clark-West statistics are still around three at one and two
quarters, and two years ahead is still nothing. The gains are slightly smaller
throughout -- `r score("u", 2, long_vintages)[["ratio"]]` against
`r score("u", 2)[["ratio"]]` at two quarters -- which is the ordinary direction
for a result re-run on data it was not found on, and a good deal better than
the alternative.

The one row that moves is four quarters, where Clark-West falls from
`r score("u", 4)[["cw"]]` to `r score("u", 4, long_vintages)[["cw"]]`. That
horizon was already the equivocal one, with the ratio and the statistic
disagreeing about how much the index was worth; the longer sample resolves the
disagreement toward the ratio's reading rather than the statistic's.

```{r oos-long-y}
t(sapply(c(1, 2, 4), score, target = "y", set = long_vintages))
```

Output growth stays unforecastable, as it was on the shorter sample.

Two limits remain, and neither is repairable with this data. There are no
vintages, so the macro variables are the revised ones throughout: a genuine
real-time user would have had worse `y`, `p` and `u` than these regressions
were given, and the paper's own real-time exercise -- which neither dataset here
can support -- is the one that settles how much that matters. And this is still
one country, so the Clark-West statistics are indicative rather than decisive
however many quarters they are computed over. See Koop and Korobilis (2014,
sec. 5) for the fuller treatment.

## References

Barigozzi, M., & Lissona, C. EA-MD-QD: Large euro area and euro member countries
datasets for macroeconomic research.
<https://github.com/BarigozziMatteo/EA-MD-QD>

Caldara, D., & Iacoviello, M. (2022). Measuring geopolitical risk. *American
Economic Review, 112*(4), 1194--1225.

Clark, T. E., & West, K. D. (2007). Approximately normal tests for equal
predictive accuracy in nested models. *Journal of Econometrics, 138*(1),
291--311.

Koop, G., & Korobilis, D. (2014). A new index of financial conditions. *European
Economic Review, 71*, 101--116.
