---
title: "Financial conditions in the euro area and its member countries"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Financial conditions in the euro area and its member countries}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

`vignette("reproducing-koop-korobilis")` runs the two exercises of Koop and
Korobilis (2014) on the US data the paper published. This one runs the same
model eleven times: once for the euro area as a whole and once for each of its
ten largest member countries, from one estimator, one block definition and one
sign convention, so that the eleven indices can be read against each other.

Two things differ from the paper, and both follow from the data rather than
from the model. The sample is short -- 103 quarters from 2000, against the
paper's 175, because that is where harmonised euro-area statistics start. And
there are no vintages, so the paper's real-time exercise cannot be repeated;
what replaces it below is weaker and is labelled as such.

```{r setup}
library("fincond")

set.seed(1234567)

data("eamd")
```

## The data

`eamd` holds one financial and one macroeconomic block for each of eleven
regions; see `?eamd`. The macro block is everywhere the `(y, p, u)` triple of
`koop$macro` -- output growth, inflation, the unemployment rate -- over an
identical 2000 Q2 to 2025 Q4 span.

```{r regions}
names(eamd)
sapply(eamd, function(z) c(financial = ncol(z$financial), obs = nrow(z$macro)))
```

The financial block is where the substantive choices are, and there are three.

It is **transformed, not shipped in levels.** EA-MD-QD publishes a
transformation code per series, and `data-raw/eamd.R` applies one: growth rates
for stocks and prices, first differences for rates, levels for survey balances
and spreads. A factor extracted from trending levels is a trend.

It is **curated, not exhaustive.** EA-MD-QD's financial class is mostly sectoral
balance sheets -- 26 of them for the euro area -- and a factor of all of it is a
factor of sectoral balance sheets. What is kept mirrors the composition of the
paper's own 18 series.

And it is **one definition applied to all eleven regions**, so that a column
name means the same thing everywhere. Four regions depart from it:

```{r columns}
core <- colnames(eamd$ea$financial)

data.frame(
  missing = sapply(eamd, function(z) paste(setdiff(core, colnames(z$financial)),
                                           collapse = " ")),
  extra = sapply(eamd, function(z) paste(setdiff(colnames(z$financial), core),
                                         collapse = " "))
)
```

Ireland publishes no maturity split of household or corporate loans and Greece
no deposit series, so those columns are absent rather than empty. Austria and
Germany carry four more than the others, because AustriaMacroData publishes a
national panel for those two and neither source has the equivalents elsewhere;
they are appended last, so subsetting to `core` recovers the strictly
comparable set. Note also how much of the block is not country-specific.
`irate_short`, `spread_money` and `usd` are the same series in all eleven -- the
money market is a single euro-area market, and so is the euro's dollar rate --
and so are the two uncertainty measures, `vix` and `gpr`, which come from
outside EA-MD-QD entirely: the CBOE volatility index from FRED, which is
`koop`'s own series and the one input here with a direct counterpart in the
paper rather than a substitute, and the global geopolitical risk index of
Caldara and Iacoviello (2022). Both are global measures, so either can move any
of these indices without anything having happened in that economy. Of the
common columns only `irate_long`, and through it `spread_term`, is genuinely
national.

## Eleven indices

```{r fit}
models <- lapply(eamd, function(d) {
  model <- create_fcimodel(y = d$macro, x = d$financial, p = 4, n = 1)
  model <- add_priors(model)
  add_posterior_coefficients(model)
})

index <- lapply(models, function(m) -fci(m, sign_on = "esi"))

all <- do.call(ts.intersect, index)
colnames(all) <- names(index)
```

The orientation deserves a section, because eleven regions turn it from a
formality into a real problem. The sign of a factor is not identified, so it has
to be fixed by convention, and `sign_on` fixes it by naming a series whose
loading should come out positive. The paper names the VIX, which loads hugely on
its factor -- and since `vix` is in this block too, `sign_on = "vix"` is the
obvious thing to reach for. It is also wrong here, and an anchor has to load
strongly, since the whole orientation turns on the sign of one number. In
magnitude:

```{r anchor}
anchor <- sapply(models, function(m) {
  l <- colMeans(factor_loadings(m))
  round(abs(l[c("vix", "gpr", "esi")]), 2)
})
anchor

apply(anchor, 1, min)
```

The VIX loads 0.06 in Italy and 0.21 in France -- small enough that the sign of
an entire country's index would turn on noise. Worse, in Greece, Italy and
Portugal it loads with the opposite sign to everywhere else, which is not a
defect in the series but a fact about those three: their factor is dominated by
a sovereign debt crisis that US equity volatility never tracked. Signing on the
VIX therefore turns those three indices upside down, and only those three:

```{r anchor-vix}
sapply(names(models), function(r) {
  round(cor(as.numeric(fci(models[[r]], sign_on = "vix")),
            as.numeric(-fci(models[[r]], sign_on = "esi"))), 2)
})
```

`gpr` is weak too, at 0.34 in Spain. What loads strongly in every region is the
survey balances -- `esi` at no less than 0.82 -- but they rise with
*loosening*, so signing on `esi` gives a high-means-loose index and the minus in
`-fci(model, sign_on = "esi")` turns it back into the paper's
high-means-stressed convention. That is the anchor used throughout below.

Eleven series on one pair of axes is unreadable, so each is drawn against the
euro area aggregate instead -- grey behind, the region itself in front:

```{r plot-all, fig.height = 8}
opar <- par(mfrow = c(4, 3), mar = c(2, 3, 2, 1), mgp = c(2, 0.6, 0))

for (r in colnames(all)) {
  ts.plot(all[, c("ea", r)], col = c("grey70", "firebrick"),
          lwd = c(1.5, 1.8), ylim = range(all), ylab = "", xlab = "",
          main = toupper(r))
  abline(h = 0, lty = "dashed", col = "grey40")
}

par(opar)
```

## Where and when the crisis was

The paper's claim about its own index is about the ranking of episodes rather
than about any one value. With eleven regions the interesting version of that
claim is not each ranking on its own but whether they differ in the way the
euro area's history says they should.

```{r peaks}
as_quarter <- function(x) {
  yr <- floor(x + 1e-8)
  paste0(yr, " Q", round((x - yr) * 4) + 1)
}

peak_of <- function(z) as_quarter(as.numeric(time(z))[which.max(z)])

data.frame(peak = sapply(index, peak_of),
           value = round(sapply(index, max), 2),
           cor_with_ea = round(apply(all, 2,
                                     function(z) cor(z, all[, "ea"])), 2))
```

They do. Eight of the eleven peak in the global financial crisis, the euro
area aggregate in 2008 Q4 -- the paper's own quarter, on entirely different
data -- and the rest a quarter or two later. Greece, Italy and Portugal peak in
2012, at the sovereign debt crisis instead, which is the single fact about the
euro area's decade that any index of its financial conditions has to reproduce,
and no part of the estimation was told about it. Ireland's peak is the highest
of the eleven, which is the right answer for the country whose banking system
failed most completely.

The episode averages say the same thing in more detail:

```{r episodes}
episode <- function(from, to) {
  round(colMeans(window(all, start = from, end = to)), 2)
}

t(rbind("2006-07 boom"      = episode(c(2006, 1), c(2007, 4)),
        "2008-09 crisis"    = episode(c(2008, 4), c(2009, 3)),
        "2011-13 sovereign" = episode(c(2011, 3), c(2013, 4)),
        "2020 pandemic"     = episode(c(2020, 1), c(2020, 4)),
        "2021 recovery"     = episode(c(2021, 1), c(2021, 4)),
        "2023-25"           = episode(c(2023, 1), c(2025, 4))))
```

Read down the two crisis columns and the periphery separates itself. Greece,
Italy, Portugal and -- narrowly -- Spain are the regions whose 2011--13 average
exceeds their 2008--09 one; in the other seven the order is the other way round.
Spain is the borderline case in both directions, and reads as one: its two
highest quarters are 2009 Q1 and 2008 Q4, so it peaks with the core, but its
2011--13 average is fractionally the higher of the two and 2012 Q2 comes third.
Spain had both crises.

The pandemic is a smaller event than 2008 in ten of the eleven, despite being
the larger shock to output. That is the index working rather than failing --
2020 was a real shock met immediately by policy support, and conditions never
tightened the way they do when the shock starts inside the financial system. The
exception is Greece, and not because 2020 was severe there but because its
2008--09 was not; the next section takes that up.

## Germany and Greece are the loose ends

Two regions sit well below the rest in agreement with the aggregate, for
opposite reasons.

```{r germany}
round(window(all[, c("de", "ea")], start = c(2002, 1), end = c(2004, 4)), 2)
round(window(all[, c("de", "ea")], start = c(2020, 1), end = c(2020, 4)), 2)
```

Germany correlates 0.40. It peaks with everyone else in 2009, but it is elevated
through 2002--04 when the aggregate is not, and it is the one region that reads
2020 as *loose* rather than tight. The first of those is defensible: the DAX had
fallen by around two thirds from its 2000 high, several large banks were in
distress, and the country was being called the sick man of Europe. The second is
the reading to treat with most caution.

Greece correlates 0.42, and its problem is the sample rather than the model. Its
2008--09 average is 0.36, the mildest of the eleven, because in 2009 the Greek
crisis had not yet begun; almost everything its factor has to work with happened
after 2010. A single factor per country is what the model is, and where a
country's cycle does not line up with the union's, that factor has no obligation
to.

## Factor loadings

Each series has its own time varying exposure to the index. The euro area's,
which `plot()` produces for any of the eleven:

```{r loadings-ea, fig.height = 7}
plot(models$ea, which = "loadings")
```

Across regions, the compact version is the average loading of each series,
signed the way the index is. What matters is not any single number but whether
the eleven columns agree:

```{r loadings-table}
# fci() flips the factor by the sign of the anchor's mean loading, and
# factor_loadings() returns the unflipped ones; the index is then negated. So
# the loading in the index's own orientation is -sign(l_esi) * l.
loadings <- sapply(models, function(m) {
  l <- colMeans(factor_loadings(m))
  -sign(l[["esi"]]) * l[core]
})
round(loadings, 1)
```

```{r loadings-signs}
counts <- data.frame(positive = rowSums(loadings > 0, na.rm = TRUE),
                     negative = rowSums(loadings < 0, na.rm = TRUE),
                     absent = rowSums(is.na(loadings)))
counts

# Series on whose sign every region that has them agrees.
unanimous <- rownames(counts)[counts$positive == 0 | counts$negative == 0]
unanimous
```

`gpr` and `spread_term` load positively in all eleven regions. Every survey
balance loads negatively in all eleven, as do household debt growth, long-term
bank lending and the short rate. In all, `r length(unanimous)` of the
`r nrow(counts)` inputs draw the same sign from every region that publishes
them.

`vix` is the informative dissenter: positive in eight regions and negative in
Greece, Italy and Portugal -- the same split that made it unusable as an anchor,
seen from the other side.

Nothing in the estimator enforces that. The sign convention fixes one series per
region, not twenty-two, and the eleven factors are estimated separately on
eleven separate blocks -- so the agreement is the indices earning a common name.

The short rate is worth a second look, since loading negatively on stress reads
backwards. It is a timing artefact rather than a puzzle: the ECB cuts when
conditions tighten, so a falling policy rate is what accompanies stress rather
than what relieves it.

The two series that genuinely split are `equity` and `spread_money`, and both
have a reason to. Equity returns in a small open market track the global cycle
more than the domestic one, so which of the two a national index picks up
varies.
And the money-market spread is a single euro-area quantity being asked to say
something country-specific, which after 2012 it cannot: it is the same series in
all eleven blocks.

## A recursive index

The paper's second exercise re-estimates the index once per real-time vintage
and shows that a reading made at the time is close to its own fully revised
counterpart. That cannot be repeated here: neither source publishes vintages.

What can be repeated is the other half of it. Re-estimating on an expanding
window answers whether the index needs data from after the quarter it is dating
-- but it uses today's revised figures throughout, so it says nothing about how
much revisions move the index. It is the weaker of the two questions and the
answer below should not be read as the paper's.

```{r recursive}
recursive_fci <- function(d, from = c(2006, 1)) {
  ends <- as.numeric(time(window(d$macro, start = from)))
  out <- numeric(length(ends))

  for (i in seq_along(ends)) {
    yr <- floor(ends[i] + 1e-8)
    qtr <- round((ends[i] - yr) * 4) + 1

    y_i <- window(d$macro, end = c(yr, qtr))
    x_i <- window(d$financial, end = c(yr, qtr))
    # Keep only the series that have started by this quarter.
    x_i <- x_i[, colSums(!is.na(x_i)) >= 2, drop = FALSE]

    m_i <- create_fcimodel(y = y_i, x = x_i, p = 4, n = 1)
    m_i <- add_priors(m_i)
    m_i <- add_posterior_coefficients(m_i)

    # Not standardised: each fit has its own sample, so its own mean and
    # variance, and rescaling every one separately would put the readings on
    # eighty different scales.
    index_i <- -fci(m_i, sign_on = "esi", standardize = FALSE)
    out[i] <- as.numeric(index_i)[length(index_i)]
  }

  ts(out, start = c(floor(ends[1] + 1e-8),
                    round((ends[1] - floor(ends[1] + 1e-8)) * 4) + 1),
     frequency = 4)
}

recursive <- lapply(eamd, recursive_fci)
```

Each entry is a one-off reading: the value the index took in that quarter,
estimated with nothing later in the sample. Against the same quarters of the
full-sample index:

```{r recursive-cor}
agreement <- sapply(names(recursive), function(r) {
  full <- -fci(models[[r]], sign_on = "esi", standardize = FALSE)
  z <- ts.intersect(recursive[[r]], full)
  cor(z[, 1], z[, 2])
})
round(sort(agreement), 3)
```

```{r recursive-plot, fig.height = 6}
compare <- function(r) {
  full <- -fci(models[[r]], sign_on = "esi", standardize = FALSE)
  z <- ts.intersect(recursive[[r]], full)
  scale_it <- function(v) (v - mean(v)) / stats::sd(v)
  ts(cbind(scale_it(z[, 1]), scale_it(z[, 2])),
     start = stats::start(z), frequency = 4)
}

opar <- par(mfrow = c(2, 2), mar = c(3, 4, 2, 1))
for (r in c("ea", "de", "el", "nl")) {
  ts.plot(compare(r), col = c("firebrick", "grey30"), lwd = c(1.8, 1),
          ylab = "Index", xlab = "", main = toupper(r))
}
par(opar)
```

The spread is the part worth keeping: from 0.73 in Greece to 0.98 in Belgium,
with a median around 0.90. It is tempting to read that against the
0.94 the genuinely real-time comparison in
`vignette("reproducing-koop-korobilis")` returns, and worth resisting -- these
exercises answer different questions, this one having been spared revisions
entirely. What is fair to say is that being spared them does not by itself make
the answer high: six of the eleven regions here still come in below 0.94.

Where agreement is weakest is the informative part. Greece is at the bottom for
a reason the panel has already shown: a recursive estimate run in 2010 had not
yet seen the sovereign crisis that the full-sample estimate uses to calibrate
it, so its early readings come from a model that does not know what is coming.
Spain, Portugal and Italy follow for the same reason. Germany is second from the
bottom for the different reason given above -- its factor is the least like the
others to begin with. At the top, Belgium, the Netherlands, Austria and Ireland
are the regions whose crisis fell squarely inside the recursion's own window.

## What the indices are for

An index of financial conditions should say something about the real economy the
real data have not said yet. Regressing each macro variable on its own value and
the index is the crude version of that test, run here for every region at once:

```{r predictive}
predictive <- function(r, target, h) {
  d <- as.data.frame(na.omit(stats::ts.intersect(
    lead = stats::lag(eamd[[r]]$macro[, target], h),
    own = eamd[[r]]$macro[, target],
    fci = index[[r]])))
  own_only <- summary(stats::lm(lead ~ own, data = d))$r.squared
  with_fci <- summary(stats::lm(lead ~ own + fci, data = d))
  c(r2_own = round(own_only, 3),
    r2_with_fci = round(with_fci$r.squared, 3),
    coef_fci = round(stats::coef(with_fci)[3, 1], 2))
}

t(sapply(names(eamd), predictive, target = "u", h = 4))
```

A year ahead, tighter financial conditions mean higher unemployment in every one
of the eleven regions, and the index adds to what unemployment's own persistence
already explains in all of them. Two years ahead the sign holds in ten of eleven
-- Portugal is the exception, and is also the region where the index adds least
at one year.

```{r predictive-y}
t(sapply(names(eamd), predictive, target = "y", h = 1))
```

For output growth the sign reverses, as it should, since growth is the outcome
you want more of: tighter conditions now, weaker growth next quarter, in ten of
the eleven regions. The fits are small because a quarterly growth rate is close
to unforecastable at any horizon by anything, which is not a defect of the
index.

Two caveats. These are in-sample fits, not forecasts: nothing here was estimated
without the data it is scored on. And the horizons overlap, so the regressions
violate the assumptions their standard errors would need, which is why none are
reported. Read the tables as descriptions of co-movement, and see Koop and
Korobilis (2014, sec. 5) for the out-of-sample version of the question.

## References

Barigozzi, M., & Lissona, C. (2024). EA-MD-QD: Large Euro Area and Euro Member
Countries Datasets for Macroeconomic Research. Zenodo.
<https://doi.org/10.5281/zenodo.10514667>

Caldara, D., & Iacoviello, M. (2022). Measuring geopolitical risk. *American
Economic Review, 112*(4), 1194--1225.

Koop, G., & Korobilis, D. (2014). A new index of financial conditions. *European
Economic Review, 71*, 101--116.
