---
title: "Reproducing Koop and Korobilis (2014)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reproducing Koop and Korobilis (2014)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

This package estimates the time varying parameter FAVAR of Koop and Korobilis
(2014) and returns the financial conditions index (FCI) it defines. This
vignette runs the two exercises the paper is built around, on the data it
published, and checks the two claims that follow from them: that the index
peaks in the fourth quarter of 2008, and that a version of it built only from
data available at the time -- a real-time index -- tracks the fully-revised
one closely enough to be usable for monitoring as it happened, not only in
hindsight.

```{r setup}
library("fincond")
data("koop")
```

`koop` holds the paper's 18 financial series and 96 real-time vintages of
three macroeconomic variables; see `?koop`. The vintages are what the second
exercise below runs over. The first uses the full sample, `macro[[96]]`.

## The index

```{r fci-fit}
model <- create_fcimodel(y = koop$macro[[96]], x = koop$financial,
                          p = 4, n = 1)
model <- add_priors(model)
model <- add_posterior_coefficients(model)

index <- fci(model, sign_on = "vix")
```

`sign_on = "vix"` fixes the sign of the index so that it rises with the VIX --
a high value means stressed conditions. That orientation is a convention, not
an estimate: see `?fci`.

```{r fci-plot}
plot(index, main = "Financial conditions index", ylab = "Index", xlab = "")
abline(h = 0, lty = "dashed", col = "grey40")
```

The paper's headline claim is about the ranking of episodes, not about any one
value, so that is what to check. The eight highest readings:

```{r fci-peaks}
as_quarter <- function(x) {
  yr <- floor(x)
  qtr <- round((x - yr) * 4) + 1
  paste0(yr, " Q", qtr)
}

ord <- order(-index)[1:8]
data.frame(quarter = as_quarter(as.numeric(time(index))[ord]),
           index = round(as.numeric(index)[ord], 2))
```

The peak is `r as_quarter(as.numeric(time(index)[which.max(index)]))`, ahead
of 1982, 1974 and 1981 -- the same ordering reported in the paper and in
`README.md`.

## Factor loadings

Each financial series has its own, time varying exposure to the index.
`plot(model, which = "loadings")` reproduces the paper's plot of them; a
loading that drifts toward zero is a series the index is no longer drawing on:

```{r loadings-plot}
plot(model, which = "loadings")
```

## A real-time index

The estimator needs no posterior simulation, which is what lets it be run
recursively: once per vintage, using only the data that vintage contains, in
well under a second in total. That recursion is the paper's second exercise,
and it is what the 96 vintages in `macro` are for.

Two things change vintage to vintage. The macro block gets one quarter longer
each time, by construction. The financial block does too, informally: several
series -- the VIX among them -- start partway through the sample, so a series
with fewer than two observations over an early vintage's span has to be left
out of that vintage's model rather than passed in empty.

```{r realtime-fit}
n_vintages <- length(koop$macro)
realtime <- data.frame(time = numeric(n_vintages), index = numeric(n_vintages))

for (i in seq_len(n_vintages)) {
  y_i <- koop$macro[[i]]

  # Keep only the financial series that have started by this vintage.
  span <- window(koop$financial, start = start(y_i), end = end(y_i))
  x_i <- koop$financial[, colSums(!is.na(span)) >= 2, drop = FALSE]

  m_i <- create_fcimodel(y = y_i, x = x_i, p = 4, n = 1)
  m_i <- add_priors(m_i)
  m_i <- add_posterior_coefficients(m_i)

  # sp500 spans the whole sample, unlike vix, so it fixes the sign here.
  index_i <- fci(m_i, sign_on = "sp500")
  realtime$time[i] <- end(index_i)[1] + (end(index_i)[2] - 1) / 4
  realtime$index[i] <- as.numeric(index_i)[length(index_i)]
}
```

Each entry of `realtime$index` is a one-off reading: the value the index would
have taken in that quarter, estimated with nothing later than that quarter's
data. Lining those up against the same quarters of the full-sample index --
built with hindsight, and re-signed on `sp500` for comparability -- shows how
much the revisions matter:

```{r realtime-compare}
index_full <- fci(model, sign_on = "sp500")
final <- as.numeric(index_full)[match(round(realtime$time, 4),
                                       round(as.numeric(time(index_full)), 4))]

start_yr <- floor(realtime$time[1])
start_qtr <- round((realtime$time[1] - start_yr) * 4) + 1
compared <- ts(cbind(realtime$index, final),
               start = c(start_yr, start_qtr), frequency = 4)

ts.plot(compared, col = c("firebrick", "grey30"), lwd = c(1.5, 1),
        ylab = "Index", xlab = "",
        main = "Real-time vs. full-sample index")
legend("topleft", legend = c("Real time", "Full sample"),
       col = c("firebrick", "grey30"), lwd = c(1.5, 1), bty = "n")
```

```{r realtime-cor}
cor(realtime$index, final)
```

A correlation above `r round(cor(realtime$index, final), 2)` between a reading
made in real time and its own fully-revised counterpart is the case for the
index as a monitoring tool: a user watching it as it was published would have
seen essentially the same signal as one looking back at it today.
