---
title: "Testing an index on your own country"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Testing an index on your own country}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

`vignette("euro-area-and-member-countries")` estimates eleven indices under one
definition and asks what they say. `vignette("austria-deep-dive")` takes one of
them apart and asks whether it forecasts. This vignette is that second question
written as a procedure, so that it can be run somewhere else -- on another of
the eleven, or on a country this package carries no data for at all.

Nothing below is specific to Austria. Every function takes a dataset and
returns a number, and the dataset is whatever you assemble. What that has to
be is two quarterly blocks:

* **`macro`**, a `ts` matrix of output growth, inflation and the unemployment
  rate, complete over the sample. This is the state of the model, not an input
  the factor competes with, so it cannot have gaps.
* **`financial`**, a `ts` matrix of whatever your country publishes. Missing
  values are allowed and expected: `create_fcimodel()` drops a series over the
  periods it lacks rather than imputing, so a series that starts halfway
  through contributes from the date it starts.

If you read one section of this vignette, read *Choosing what to forecast
against*. It is the choice that decides whether the rest of the exercise means
anything.

```{r setup}
library("fincond")

set.seed(1234567)

data("eamd")
```

## The workflow in three functions

The exercise is a recursion, a set of competing forecasts at each step of it,
and a score. Copy these three functions; nothing in them names a country.

The first re-estimates the index once per origin, keeping the whole filtered
path rather than only its last value, because the forecasting step needs the
index history each vintage would have had.

```{r vintages}
vintages_of <- function(d, from, n = 1, anchor = "esi") {
  origins <- as.numeric(time(window(d$macro, start = from)))

  paths <- lapply(origins, function(o) {
    yr <- floor(o + 1e-8)
    qtr <- round((o - yr) * 4) + 1

    x_o <- window(d$financial, end = c(yr, qtr))
    # Keep only the series that have started by this quarter.
    x_o <- x_o[, colSums(!is.na(x_o)) >= 2, drop = FALSE]

    m_o <- create_fcimodel(y = window(d$macro, end = c(yr, qtr)), x = x_o,
                           p = 4, n = n)
    m_o <- add_posterior_coefficients(add_priors(m_o))

    # Not standardised: each fit has its own sample, so its own mean and
    # variance, and rescaling every one separately would put the readings on
    # as many different scales as there are origins.
    fci(m_o, sign_on = anchor, standardize = FALSE)
  })

  list(origins = origins, vintage = paths, macro = d$macro)
}

last_of <- function(z) as.numeric(z)[length(z)]
```

The second makes four forecasts at each origin: three benchmarks and the
index. The rule is that nothing estimated at origin `t` may touch data after
`t`, which covers the regression coefficients as well as the index.

```{r forecasts}
forecasts <- function(target, h, set, cumulative = TRUE) {
  y <- set$macro[, target]
  stamp <- as.numeric(time(y))

  rows <- lapply(seq_along(set$origins), function(k) {
    path <- set$vintage[[k]]

    ahead <- vapply(seq_len(h), function(j) {
      hit <- which(abs(stamp - (set$origins[k] + j / 4)) < 1e-6)
      if (length(hit) == 0) NA_integer_ else hit[1]
    }, integer(1))
    if (anyNA(ahead)) return(NULL)

    realised <- if (cumulative) {
      mean(as.numeric(y)[ahead])
    } else {
      as.numeric(y)[ahead[h]]
    }

    seen <- window(y, end = end(path))
    lead <- stats::lag(seen, h)
    if (cumulative) {
      lead <- stats::filter(lead, rep(1 / h, h), sides = 1)
    }

    d <- as.data.frame(na.omit(stats::ts.intersect(
      lead = lead, own = seen, fci = path)))
    if (nrow(d) < 12) return(NULL)

    now <- data.frame(own = last_of(seen), fci = last_of(path))
    predict_with <- function(f) {
      as.numeric(stats::predict(stats::lm(f, data = d), newdata = now))
    }

    data.frame(origin = set$origins[k], actual = realised,
               mean = mean(d$lead),          # recursive mean of the target
               rw = last_of(seen),           # random walk
               ar = predict_with(lead ~ own),
               fci = predict_with(lead ~ own + fci))
  })

  do.call(rbind, rows)
}
```

The third scores the index against whichever benchmark did best. The ratio of
root mean squared errors is what a user would have experienced; the Clark-West
statistic asks whether the index belongs in the population regression,
correcting for the estimation noise a plain Diebold-Mariano test charges the
larger model twice for. All three benchmarks are nested in the regression with
the index -- the mean sets both slopes to zero, the random walk fixes the
own-lag coefficient at one -- so the correction applies to each.

```{r score}
# Long-run variance, since h-step forecasts overlap for h > 1.
hac <- function(z, lag) {
  n <- length(z)
  z <- z - mean(z)
  v <- sum(z^2) / n
  for (l in seq_len(lag)) {
    v <- v + 2 * (1 - l / (lag + 1)) * sum(z[(l + 1):n] * z[1:(n - l)]) / n
  }
  v
}

score <- function(target, h, set, cumulative = TRUE) {
  f <- forecasts(target, h, set, cumulative)
  rmse <- function(col) sqrt(mean((f$actual - f[[col]])^2))

  competing <- vapply(c("mean", "rw", "ar"), rmse, numeric(1))
  best <- names(competing)[which.min(competing)]

  e_bench <- f$actual - f[[best]]
  e_fci <- f$actual - f$fci
  adjusted <- e_bench^2 - (e_fci^2 - (f[[best]] - f$fci)^2)

  list(best = best,
       ratio = rmse("fci") / min(competing),
       ratio_ar = rmse("fci") / competing[["ar"]],
       cw = mean(adjusted) /
         sqrt(hac(adjusted, max(h - 1, 0)) / length(adjusted)))
}
```

## Choosing what to forecast against

An own-lag regression is the obvious benchmark for a persistent series and a
surprisingly weak one. An unemployment rate is close to a unit root; estimating
its autoregressive coefficient on eighty quarters shrinks it below one and
pulls the forecast toward a sample mean the level has drifted away from. Two
alternatives cost nothing to compute. The recursive mean is what a forecaster
with no model at all would say. The random walk imposes the unit root the
regression is trying to estimate.

Build the recursions once and score off them as many times as you like:

```{r eleven}
one_factor <- lapply(eamd, vintages_of, from = c(2006, 1), n = 1)
two_factor <- lapply(eamd, vintages_of, from = c(2006, 1), n = 2)

horizons <- c(1, 2, 4, 8)

picked <- t(sapply(two_factor, function(s) {
  vapply(horizons, function(h) score("u", h, s)$best, character(1))
}))
colnames(picked) <- paste0("h=", horizons)
picked
```

The random walk is the benchmark to beat almost everywhere. Austria is the
single exception, and only up to a year: Austrian unemployment mean-reverts
where most of the euro area's drifts. That is worth knowing when reading
`vignette("austria-deep-dive")`, which is built on the one country where the
habitual benchmark happens to be the right one.

It matters more for anyone scoring against an own-lag regression alone. Here is
what that choice is worth at two years, as the ratio of the index's error to
the best benchmark and to the own-lag regression:

```{r benchmark-cost}
cost <- t(sapply(two_factor, function(s) {
  z <- score("u", 8, s)
  c(vs_best = z$ratio, vs_own_lag = z$ratio_ar)
}))
round(cost, 3)
```

Scored against the own-lag regression the index appears to win two years ahead
in ten of eleven regions. Scored against the better of a random walk and a
recursive mean it loses in all eleven. The first column is the honest one, and
nothing about the index changed between them.

## What the index is actually worth

```{r eleven-table}
summarise <- function(sets, cumulative = TRUE) {
  t(sapply(horizons, function(h) {
    z <- lapply(sets, function(s) score("u", h, s, cumulative))
    r <- vapply(z, `[[`, numeric(1), "ratio")
    c(h = h, wins = sum(r < 1), mean_ratio = round(mean(r), 3),
      cw_sig = sum(vapply(z, `[[`, numeric(1), "cw") > 1.645))
  }))
}

summarise(two_factor)
```

The index beats the best simple benchmark at one and two quarters, in ten and
nine of the eleven regions, with Clark-West clearing the one-sided five per
cent value in eleven and ten. At a year it is close to a coin flip. At two
years it loses everywhere.

That is the result. A financial conditions index carries real information about
unemployment one and two quarters out, across eleven countries, and the
long-horizon predictive content that an own-lag benchmark appears to show is an
artefact of the benchmark.

```{r eleven-detail}
detail <- t(sapply(two_factor, function(s) {
  vapply(horizons, function(h) score("u", h, s)$ratio, numeric(1))
}))
colnames(detail) <- paste0("h=", horizons)
round(detail, 3)
```

The dispersion is the other half of the finding. The Netherlands and Ireland
give up around a fifth of the benchmark's error at one quarter; Portugal never
beats it at any horizon. Show all eleven rather than a favourable one.

## The two remaining choices

**What to forecast.** `cumulative = TRUE` takes the target to be the average of
the variable over `t + 1` to `t + h` rather than the single quarter `t + h`.
Conditions tighten over a stretch of quarters and unemployment answers over a
stretch of quarters, so asking the index to name one particular quarter loads
the exercise with movement no financial block claims to see.

**How many factors.** `n = 2` estimates a second factor that is never used as a
predictor; the regressions take the first factor and nothing else. What it does
is give the variation it loads on somewhere else to go.

```{r choices}
wins <- rbind(
  "endpoint, n = 1"   = summarise(one_factor, cumulative = FALSE)[, "wins"],
  "cumulative, n = 1" = summarise(one_factor)[, "wins"],
  "cumulative, n = 2" = summarise(two_factor)[, "wins"])
colnames(wins) <- paste0("h=", horizons)
wins
```

The cumulative target earns its place at a year, where it roughly doubles the
count of regions in which the index beats the best benchmark. It changes
nothing at one quarter, where the two targets are identical by construction,
and it cannot rescue two years, where nothing does. The second factor is worth
little on this evidence and is not worth assuming; check it on your own data
and be prepared for the answer to be one.

## What not to spend time on

Most of the search behind these numbers found nothing. Those negatives are
worth more to a reader starting out than the settings that worked.

* **Tuning the estimator.** Lag order `p` in `{1, 2, 4}`, all four forgetting
  factor settings in `?add_priors`, and dropping the national block: every
  resulting ratio sat within a hundredth of the baseline.
* **Selecting on the likelihood.** `posterior$loglik` does not track
  out-of-sample accuracy.
* **Widening the autoregression.** AR(2) and AR(4) benchmarks are *worse* than
  AR(1), not better, which is a symptom of the same disease: more coefficients
  estimated on a near-unit-root series in a short sample. Adding lags is not a
  substitute for imposing the unit root.
* **More regressors in the second stage.** Adding the other two macro variables
  to the predictive regression took Austria's one quarter ratio from 0.958 to
  1.170. Eighty observations do not support it.
* **Rolling instead of expanding windows.** Worse at every horizon tried.
* **Pooling the slope across countries.** A panel estimate with country fixed
  effects, computed in real time: a fifth of a point at one and two quarters,
  clearly worse at four and eight.
* **A longer warm-up before scoring.** Requiring 40 observations before an
  origin counts improves Austria's two year ratio against the own-lag
  regression from 1.048 to 0.996 -- and scoring the short warm-up on that same
  restricted set of origins gives 0.996 as well. The whole gain was dropping
  the hard early origins. If you change the scoring sample, compare on the
  sample, not on the headline.

## Adding a national series

The part of the workflow that genuinely is country-specific is the block. Euro
area members have an advantage here, because the ECB publishes a Bank Lending
Survey for each of them, and credit standards are the one thing Koop and
Korobilis (2014) had that `eamd` does not: a measure of whether banks will lend
at all, as against the price of credit or the stock of it outstanding.

The Data Portal API takes a country code, so this is one string away from being
your country's series. It is not run when this vignette is built, since that
would need a network connection:

```{r bls-fetch, eval = FALSE}
# Credit standards on loans to enterprises, net percentage reporting a
# tightening. Replace "AT" with your own two-letter code; "U2" is the euro area.
bls_url <- function(country, key = "O.E.Z.B3.ST.S.FNET") {
  paste0("https://data-api.ecb.europa.eu/service/data/BLS/Q.",
         country, ".ALL.", key, "?format=csvdata&startPeriod=2000-Q1")
}

d <- read.csv(bls_url("AT"))
d <- d[order(d$TIME_PERIOD), ]
standards <- ts(d$OBS_VALUE,
                start = c(as.integer(substr(d$TIME_PERIOD[1], 1, 4)),
                          as.integer(substr(d$TIME_PERIOD[1], 7, 7))),
                frequency = 4)
```

Appending it needs one piece of care, because `cbind()` on a `ts` matrix
prefixes the existing column names, which would rename the anchor `esi` out
from under `fci()`:

```{r bls-augment, eval = FALSE}
augment <- function(block, extra) {
  stamp <- round(as.numeric(stats::time(block)), 6)
  add <- matrix(NA_real_, nrow(block), length(extra),
                dimnames = list(NULL, names(extra)))
  for (nm in names(extra)) {
    s <- extra[[nm]]
    pos <- match(round(as.numeric(stats::time(s)), 6), stamp)
    add[pos[!is.na(pos)], nm] <- as.numeric(s)[!is.na(pos)]
  }
  out <- cbind(as.matrix(block), add)
  colnames(out) <- c(colnames(block), names(extra))
  stats::ts(out, start = stats::start(block),
            frequency = stats::frequency(block))
}

richer <- eamd$at
richer$financial <- augment(richer$financial, list(std_nfc = standards))
score("u", 4, vintages_of(richer, from = c(2006, 1), n = 2))
```

The series is right. It runs from 2003 Q1, its highest reading is 2009 Q1 at
100 -- every bank in the panel tightening at once -- and that is the quarter
Austria's index dates as its trough. The effect on the forecasts is almost
nothing: credit standards for enterprises move Austria's four quarter ratio by
two thousandths, and the forward-looking version by four.

That is worth stating plainly, because it was the best-motivated candidate in
the whole exercise. Twenty additions to the Austrian block were tried --
government balance sheets, short-term corporate credit, economy-wide loans and
deposits, four more confidence surveys, household mortgage volumes, all four
Bank Lending Survey standards series and loan demand -- and the largest
improvement at any horizon from any of them was four thousandths.

The lesson is not that credit standards are uninformative. It is that a
twenty-six series block summarised by one factor is already saturated: the
factor is a weighted average, and a twenty-seventh correlated series changes
the average very little while adding noise to its estimation. If your own block
is much smaller, the arithmetic is different and a well chosen addition may
well pay. Test it the way this section does, and judge it at the horizons where
the index is actually earning its keep -- which, on this evidence, means one and
two quarters.

## Caveats to carry with you

**The benchmark set is still three simple rules.** A forecaster with a properly
specified time series model of your country's unemployment would be harder to
beat than any of them. Everything above says the index beats a random walk at
short horizons, which is a real but modest claim.

**The sample is thin.** Eighty origins of one country is not much, and if you
select a specification on those eighty you have spent them. The eleven-region
table is the form the check takes here; a second, longer dataset is the other,
and `vignette("austria-deep-dive")` uses both.

**These are not real-time data.** `eamd` has no vintages, so the macro
variables are the revised figures throughout. Where your country publishes
vintages, use them: it is the one improvement here that would make the exercise
strictly more honest rather than merely different.

**Eleven countries are not eleven observations.** They share a monetary policy
and much of a business cycle, so the counts above describe a correlated panel
rather than eleven independent tests.

## References

Clark, T. E., & West, K. D. (2007). Approximately normal tests for equal
predictive accuracy in nested models. *Journal of Econometrics, 138*(1),
291--311.

European Central Bank. Bank Lending Survey.
<https://data.ecb.europa.eu/data/datasets/BLS>

Koop, G., & Korobilis, D. (2014). A new index of financial conditions. *European
Economic Review, 71*, 101--116.
