---
title: "Testing an index on your own country"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Testing an index on your own country}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

`vignette("euro-area-and-member-countries")` estimates eleven indices under one
definition and asks what they say. `vignette("austria-deep-dive")` takes one of
them apart and asks whether it forecasts. This vignette is that second question
written as a procedure, so that it can be run somewhere else -- on another of
the eleven, or on a country this package carries no data for at all.

Nothing below is specific to Austria. Every function takes a dataset and
returns a number, and the dataset is whatever you assemble. What that has to
be is two quarterly blocks:

* **`macro`**, a `ts` matrix of output growth, inflation and the unemployment
  rate, complete over the sample. This is the state of the model, not an input
  the factor competes with, so it cannot have gaps.
* **`financial`**, a `ts` matrix of whatever your country publishes. Missing
  values are allowed and expected: `create_fcimodel()` drops a series over the
  periods it lacks rather than imputing, so a series that starts halfway
  through contributes from the date it starts.

```{r setup}
library("fincond")

set.seed(1234567)

data("eamd")
```

## The workflow in three functions

The exercise is a recursion, a pair of forecasts at each step of it, and a
score. Copy these three functions; nothing in them names a country.

The first re-estimates the index once per origin, keeping the whole filtered
path rather than only its last value, because the forecasting step needs the
index history each vintage would have had.

```{r vintages}
vintages_of <- function(d, from, n = 1, anchor = "esi") {
  origins <- as.numeric(time(window(d$macro, start = from)))

  paths <- lapply(origins, function(o) {
    yr <- floor(o + 1e-8)
    qtr <- round((o - yr) * 4) + 1

    x_o <- window(d$financial, end = c(yr, qtr))
    # Keep only the series that have started by this quarter.
    x_o <- x_o[, colSums(!is.na(x_o)) >= 2, drop = FALSE]

    m_o <- create_fcimodel(y = window(d$macro, end = c(yr, qtr)), x = x_o,
                           p = 4, n = n)
    m_o <- add_posterior_coefficients(add_priors(m_o))

    # Not standardised: each fit has its own sample, so its own mean and
    # variance, and rescaling every one separately would put the readings on
    # as many different scales as there are origins.
    fci(m_o, sign_on = anchor, standardize = FALSE)
  })

  list(origins = origins, vintage = paths, macro = d$macro)
}

last_of <- function(z) as.numeric(z)[length(z)]
```

The second makes two forecasts at each origin: a benchmark on the target's own
lag, and the same regression with the index added. The rule is that nothing
estimated at origin `t` may touch data after `t`, which covers the regression
coefficients as well as the index.

```{r forecasts}
forecasts <- function(target, h, set, cumulative = TRUE) {
  y <- set$macro[, target]
  stamp <- as.numeric(time(y))

  rows <- lapply(seq_along(set$origins), function(k) {
    path <- set$vintage[[k]]

    ahead <- vapply(seq_len(h), function(j) {
      hit <- which(abs(stamp - (set$origins[k] + j / 4)) < 1e-6)
      if (length(hit) == 0) NA_integer_ else hit[1]
    }, integer(1))
    if (anyNA(ahead)) return(NULL)

    realised <- if (cumulative) {
      mean(as.numeric(y)[ahead])
    } else {
      as.numeric(y)[ahead[h]]
    }

    seen <- window(y, end = end(path))
    lead <- stats::lag(seen, h)
    if (cumulative) {
      lead <- stats::filter(lead, rep(1 / h, h), sides = 1)
    }

    d <- as.data.frame(na.omit(stats::ts.intersect(
      lead = lead, own = seen, fci = path)))
    if (nrow(d) < 12) return(NULL)

    now <- data.frame(own = last_of(seen), fci = last_of(path))
    predict_with <- function(f) {
      as.numeric(stats::predict(stats::lm(f, data = d), newdata = now))
    }

    data.frame(origin = set$origins[k], actual = realised,
               bench = predict_with(lead ~ own),
               fci = predict_with(lead ~ own + fci))
  })

  do.call(rbind, rows)
}
```

The third scores them. The ratio of root mean squared errors is what a user
would have experienced. The Clark-West statistic asks whether the index belongs
in the population regression, correcting for the estimation noise that a
plain Diebold-Mariano test charges the larger model twice for.

```{r score}
# Long-run variance, since h-step forecasts overlap for h > 1.
hac <- function(z, lag) {
  n <- length(z)
  z <- z - mean(z)
  v <- sum(z^2) / n
  for (l in seq_len(lag)) {
    v <- v + 2 * (1 - l / (lag + 1)) * sum(z[(l + 1):n] * z[1:(n - l)]) / n
  }
  v
}

score <- function(target, h, set, cumulative = TRUE) {
  f <- forecasts(target, h, set, cumulative)
  e_bench <- f$actual - f$bench
  e_fci <- f$actual - f$fci

  # Clark and West (2007): under the null the larger model estimates a
  # coefficient that is truly zero, and the noise in doing so inflates its
  # squared error. The last term subtracts exactly that penalty.
  adjusted <- e_bench^2 - (e_fci^2 - (f$bench - f$fci)^2)

  c(ratio = sqrt(mean(e_fci^2)) / sqrt(mean(e_bench^2)),
    cw = mean(adjusted) / sqrt(hac(adjusted, max(h - 1, 0)) / length(adjusted)))
}
```

## Running it on eleven countries

The recursion is the expensive part, so build it once per specification and
score off it as many times as you like. This is the whole exercise for all
eleven regions, at two factors, on an expanding window from 2006 Q1:

```{r eleven}
one_factor <- lapply(eamd, vintages_of, from = c(2006, 1), n = 1)
two_factor <- lapply(eamd, vintages_of, from = c(2006, 1), n = 2)

horizons <- c(1, 2, 4, 8)

ratios <- t(sapply(two_factor, function(s) {
  vapply(horizons, function(h) score("u", h, s)[["ratio"]], numeric(1))
}))
colnames(ratios) <- paste0("h=", horizons)
round(ratios, 3)
```

```{r eleven-cw}
cw <- t(sapply(two_factor, function(s) {
  vapply(horizons, function(h) score("u", h, s)[["cw"]], numeric(1))
}))
colnames(cw) <- paste0("h=", horizons)
round(cw, 2)
```

A ratio below one is the index beating the benchmark. It does so in every
region at one, two and four quarters, and in all but one at two years. The
spread is wide -- Ireland takes nearly thirty per cent off the benchmark's root
mean squared error and the Netherlands a fifth, Portugal almost nothing -- and
the point of showing all eleven rather than a favourable one is that the spread
is the finding.

Note where Austria sits. The country `vignette("austria-deep-dive")` spends a
whole vignette on is the second *weakest* of the eleven at one quarter. Whatever
its index is worth, it is not the case that makes the method look good, which
is worth knowing before reading its results as a promise about your own
country.

Portugal is the honest failure: the only region losing at two years, and the
only one whose Clark-West statistics never clear two. A workflow that could not
produce that row would not be telling you anything.

## The two choices that matter, and which of them generalises

Two decisions sit inside `vintages_of()` and `forecasts()` above, and the
Austrian vignette settles both by trying the alternatives. Run the same test
on your own data rather than inheriting its answers, because one of the two
transfers and the other does not.

**What to forecast.** `cumulative = TRUE` takes the target to be the average of
the variable over `t + 1` to `t + h` -- the path -- rather than the single
quarter `t + h`. Conditions tighten over a stretch of quarters and
unemployment answers over a stretch of quarters, so asking the index to name
one particular quarter loads the exercise with movement no financial block
claims to see.

**How many factors.** `n = 2` estimates a second factor that is never used as a
predictor; the regressions take the first factor and nothing else. What it does
is give the variation it loads on somewhere else to go.

```{r choices}
compare <- function(h) {
  data.frame(
    single_n1 = vapply(one_factor, function(s)
      score("u", h, s, cumulative = FALSE)[["ratio"]], numeric(1)),
    cumul_n1 = vapply(one_factor, function(s)
      score("u", h, s)[["ratio"]], numeric(1)),
    cumul_n2 = vapply(two_factor, function(s)
      score("u", h, s)[["ratio"]], numeric(1)))
}

round(compare(4), 3)
```

```{r choices-count}
tally <- function(h) {
  x <- compare(h)
  c(cumulative_better = sum(x$cumul_n1 < x$single_n1),
    two_factors_better = sum(x$cumul_n2 < x$cumul_n1),
    of = nrow(x))
}

rbind("h = 4" = tally(4), "h = 8" = tally(8))
```

The cumulative target is a rule. It improves ten of the eleven at both
horizons; the exception at four quarters is Greece, by two thousandths, and at
two years it is Portugal, which is the region that fails on every other count
as well.

The second factor is not a rule. It is close to a coin flip at four quarters
and only a little better at two years, and the countries it helps are not
obviously the ones you would have picked. Treat `n` as something to check on
your own data, and be prepared for the answer to be one.

## What not to spend time on

The Austrian vignette's result came out of a search, and most of the search
found nothing. Those negatives are worth more to a reader starting out than the
one setting that worked, because each of them is a week you need not spend.

* **Tuning the estimator.** Lag order `p` in `{1, 2, 4}`, all four forgetting
  factor settings in `?add_priors`, and dropping the national block: every
  resulting ratio sat within a hundredth of the baseline. The index is
  remarkably insensitive to its own tuning parameters.
* **Selecting on the likelihood.** `posterior$loglik` does not track
  out-of-sample accuracy. On Austria the specification with the best one step
  ahead predictive likelihood was not the one with the best forecasts.
* **A richer benchmark.** Replacing the own-lag regression with an AR(2) or
  AR(4) makes the *benchmark* worse, not better, and the index still beats it.
  A weak benchmark is not what is producing these numbers.
* **More regressors in the second stage.** Adding the other two macro variables
  to the predictive regression took Austria's one quarter ratio from 0.958 to
  1.170. Eighty observations do not support it.
* **Rolling instead of expanding windows.** Worse at every horizon tried.
* **Pooling the slope across countries.** A panel estimate of the predictive
  regression, with country fixed effects, computed in real time: a fifth of a
  point at one and two quarters, clearly worse at four and eight.
* **A longer warm-up before scoring.** This one is a trap. Requiring 40
  observations before an origin counts improves Austria's two year ratio from
  1.048 to 0.996 -- and scoring the short warm-up on that same restricted set
  of origins gives 0.996 as well. The whole gain was dropping the hard early
  origins. If you change the scoring sample, compare on the sample, not on the
  headline.

## Adding a national series

The part of the workflow that genuinely is country-specific is the block. Here
the euro area members have an advantage, because the ECB publishes a Bank
Lending Survey for each of them, and credit standards are the one thing
Koop and Korobilis (2014) had that `eamd` does not: a measure of whether banks
will lend at all, as against the price of credit or the stock of it outstanding.

The Data Portal API takes a country code, so this is one string away from being
your country's series. It is not run when this vignette is built, since that
would need a network connection:

```{r bls-fetch, eval = FALSE}
# Credit standards on loans to enterprises, net percentage reporting a
# tightening. Replace "AT" with your own two-letter code; "U2" is the euro area.
bls_url <- function(country, key = "O.E.Z.B3.ST.S.FNET") {
  paste0("https://data-api.ecb.europa.eu/service/data/BLS/Q.",
         country, ".ALL.", key, "?format=csvdata&startPeriod=2000-Q1")
}

d <- read.csv(bls_url("AT"))
d <- d[order(d$TIME_PERIOD), ]
standards <- ts(d$OBS_VALUE,
                start = c(as.integer(substr(d$TIME_PERIOD[1], 1, 4)),
                          as.integer(substr(d$TIME_PERIOD[1], 7, 7))),
                frequency = 4)
```

Appending it needs one piece of care, because `cbind()` on a `ts` matrix
prefixes the existing column names, which would rename the anchor `esi` out
from under `fci()`:

```{r bls-augment, eval = FALSE}
augment <- function(block, extra) {
  stamp <- round(as.numeric(stats::time(block)), 6)
  add <- matrix(NA_real_, nrow(block), length(extra),
                dimnames = list(NULL, names(extra)))
  for (nm in names(extra)) {
    s <- extra[[nm]]
    pos <- match(round(as.numeric(stats::time(s)), 6), stamp)
    add[pos[!is.na(pos)], nm] <- as.numeric(s)[!is.na(pos)]
  }
  out <- cbind(as.matrix(block), add)
  colnames(out) <- c(colnames(block), names(extra))
  stats::ts(out, start = stats::start(block),
            frequency = stats::frequency(block))
}

richer <- eamd$at
richer$financial <- augment(richer$financial, list(std_nfc = standards))
score("u", 4, vintages_of(richer, from = c(2006, 1), n = 2))
```

The series is right. It runs from 2003 Q1, its highest reading is 2009 Q1 at
100 -- every bank in the panel tightening at once -- and that is the quarter
Austria's index dates as its trough.

The effect on the forecasts is almost nothing. Credit standards for enterprises
move the four quarter ratio from 0.891 to 0.893; the version asking banks what
they expect over the next three months, which is the more forecast-shaped
question, moves it to 0.887. Household standards for house purchase and for
consumer credit land in between, and adding loan demand alongside standards is
worse than either.

That is worth stating plainly, because it was the best-motivated candidate in
the whole exercise. Twenty additions to the Austrian block were tried --
government balance sheets, short-term corporate credit, economy-wide loans and
deposits, four more confidence surveys, household mortgage volumes, all four
Bank Lending Survey standards series and loan demand -- and the largest
improvement at any horizon from any of them was four thousandths. Several
bought a better one quarter ratio and paid for it at two years.

The lesson is not that credit standards are uninformative. It is that a
twenty-six series block summarised by one factor is already saturated: the
factor is a weighted average, and a twenty-seventh correlated series changes
the average very little while adding noise to its estimation. If your own block
is much smaller than this one, the arithmetic is different and a well chosen
addition may well pay. Test it the way this section does, and judge it at the
horizons where the index is actually earning its keep rather than at `h = 1`.

## Caveats to carry with you

Three, and none of them is repairable by trying harder.

**The sample is thin.** Eighty origins of one country is not much, and if you
select a specification on those eighty you have spent them. The Austrian
vignette checks its two choices on a second, longer Austrian dataset for
exactly this reason; the eleven-region table above is the other form the check
can take. Do one or the other before believing a number.

**These are not real-time data.** `eamd` has no vintages, so the macro
variables are the revised figures throughout. A genuine real-time user would
have had worse `y`, `p` and `u` than these regressions were given. Where your
country publishes vintages, use them: it is the one improvement here that would
make the exercise strictly more honest rather than merely different.

**One country is one observation.** A Clark-West statistic computed over eighty
overlapping forecasts of a single country is indicative. It is the eleven rows
agreeing, not any one of them, that makes the case.

## References

Clark, T. E., & West, K. D. (2007). Approximately normal tests for equal
predictive accuracy in nested models. *Journal of Econometrics, 138*(1),
291--311.

European Central Bank. Bank Lending Survey.
<https://data.ecb.europa.eu/data/datasets/BLS>

Koop, G., & Korobilis, D. (2014). A new index of financial conditions. *European
Economic Review, 71*, 101--116.
